---
title: "10. Regression"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

```{r,message=FALSE}
library(R2jags)
library(ggmcmc)
library(loo)
library(rstan)
library(here)
theme_set(theme_bw())
# to run R in parallel
options(mc.cores = parallel::detectCores())
# To keep a compiled version of the code so you don't have to recompile
rstan_options(auto_write = TRUE)
```

# Frequentist version of linear regression for coarse woody debris

```{r}
cwd<-data.frame(TreeDens=c(1270, 1210, 1800, 1875, 1300, 2150, 1330, 964, 961, 1400, 1280, 976, 771, 833, 883, 956),
  CWD=c(121, 41, 183, 130, 127, 134, 65, 52, 12, 46, 54, 97, 1, 4, 1, 4))
head(cwd)
cwdlm<-lm(CWD~TreeDens,data=cwd)
summary(cwdlm)
ggplot(cwd,aes(x=TreeDens,y=CWD))+
  geom_point()+
  stat_smooth(method="lm")
predict(cwdlm,interval="confidence")
```

# Woody debris regression, Bayesian. Box 5.1

```{r}
write("model{
  a ~ dnorm (0, 1.0E-6)  	# prior for the intercept
  b ~ dnorm (0, 1.0E-6)  		# prior for the slope
  prec ~ dgamma(0.001, 0.001)  # prior for the precision 
  for (i in 1:16)  # for each of the 16 lakes
  {
    ymean[i] <- a + b*TreeDens[i]  # the predicted y value y=a+bx
    CWD[i] ~ dnorm(ymean[i], prec) #  the actual y, normally distributed 
    pred.obs[i] ~ dnorm(ymean[i], prec) # with error added
    resid[i]<-CWD[i]-ymean[i]	
    sresid[i]<-(CWD[i]-ymean[i])*sqrt(prec) 
    sresid2[i]<-sresid[i]*sresid[i] 
    rep.sresid2[i]<-(pred.obs[i]-ymean[i])*(pred.obs[i]-ymean[i])*prec
    LL[i]<- -log(1/sqrt(prec))-0.5*log(2*3.14159)-0.5*(CWD[i]-ymean[i])^2*prec
  }
  for(j in 1:n.new) {
    mean.new[j]<-a+b*x.new[j]
    pred.new[j]~dnorm(mean.new[j],prec)
  }
  #summary statistics
  prob.pos.b<-step(b)
  sdy<-sd(CWD[])
  sdy2<-sdy*sdy
  r.squared<-1-1/(prec*sdy2)
  chi.square.obs<-sum(sresid2[])
  chi.square.rep<-sum(rep.sresid2[])
  p.value<-step(chi.square.obs-chi.square.rep)
}
",file=here("JAGS_mods", "box5.1.txt"))

data1<-list(TreeDens=cwd$TreeDens,
  CWD=cwd$CWD,
  x.new=c(800,1000,1200,1400,1600,1800,2000,2200), #New data for predicting
  n.new=8)

#inits are probably not needed
init1<-list(list(a=0, b=0, prec=100),list(a=1, b=1, prec=50))
#fit model
res1<-jags(data1,
  init1,
  parameters.to.save = c("a","b","prec",
  "ymean","pred.obs","mean.new","prob.pos.b","r.squared","LL"),
  model.file = here("JAGS_mods", "box5.1.txt"),
  n.chains=2,n.thin=10,n.iter=110000,n.burnin=10000)
dim(res1$BUGSoutput$summary)
#Print summary for only estimated parameters
round(res1$BUGSoutput$summary[c("a","b","prec"),],4)
```

## plot credible interval. 

This requires pulling out only the parameters we need from the summary table

```{r}
dimnames(res1$BUGSoutput$summary)
n<-length(data1$TreeDens)
meanrows<-paste0("ymean[",1:n,"]")
round(res1$BUGSoutput$summary[meanrows,],2)
cwd$Predict<-res1$BUGSoutput$summary[meanrows,"mean"]
cwd$lci<-res1$BUGSoutput$summary[meanrows,"2.5%"]
cwd$uci<-res1$BUGSoutput$summary[meanrows,"97.5%"]
head(cwd)
ggplot(cwd,aes(x=TreeDens,y=CWD))+geom_point()+
  geom_ribbon(aes(x=TreeDens,ymin=lci,ymax=uci),alpha=0.3)+
  geom_line(aes(x=TreeDens,y=Predict))+
  xlab("Tree density")+ylab("Course woody debris")
```

## plot prediction interval

```{r}
predrows<-paste0("pred.obs[",1:length(data1$TreeDens),"]")
cwd$lpi<-res1$BUGSoutput$summary[predrows,"2.5%"]
cwd$upi<-res1$BUGSoutput$summary[predrows,"97.5%"]
ggplot(cwd,aes(x=TreeDens,y=CWD))+geom_point()+
  geom_ribbon(aes(x=TreeDens,ymin=lpi,ymax=upi),alpha=0.3)+
  geom_ribbon(aes(x=TreeDens,ymin=lci,ymax=uci),alpha=0.5)+
  geom_line(aes(x=TreeDens,y=Predict))+
  xlab("Tree density")+ylab("Course woody debris")
```

## Plot credible interval of predictions from new data x.new 

```{r}
newtree<-data.frame(TreeDens=data1$x.new)
meanrowsnew<-paste0("mean.new[",1:length(data1$x.new),"]")
newtree$Predict<-res1$BUGSoutput$summary[meanrowsnew,"mean"]
newtree$lci<-res1$BUGSoutput$summary[meanrowsnew,"2.5%"]
newtree$uci<-res1$BUGSoutput$summary[meanrowsnew,"97.5%"]
ggplot(newtree)+geom_point(data=cwd,aes(x=TreeDens,y=CWD))+
  geom_ribbon(aes(x=TreeDens,ymin=lci,ymax=uci),alpha=0.3)+
  geom_line(aes(x=TreeDens,y=Predict))+
  xlab("Tree density")+ylab("Course woody debris")
```

## Summary statistics

```{r}
round(res1$BUGSoutput$summary[c("a","b","prec","prob.pos.b","r.squared"),],4)
```


## Get checking stats

```{r}
params.cwd=c("a","b","prec","p.value","resid","pred.obs","ymean",
  "sresid","prob.pos.b","r.squared","chi.square.obs","LL",
  "chi.square.rep","mean.new","pred.new")

res1check<-jags(data=data1,
  inits=init1,
  parameters.to.save=params.cwd,
  model.file=here("JAGS_mods","Box5.1.txt"),
  n.chains=2,n.iter=110000,n.burnin=10000,n.thin=20)
res1check$BUGSoutput$summary[c("a","b","prec"),]
```

# To make model checking and diagnostic plots

```{r}
n<-length(data1$TreeDens)
residrows<-paste0("resid[",1:n,"]")
sresidrows<-paste0("sresid[",1:n,"]")
meanrows<-paste0("ymean[",1:n,"]")
sumtab<-res1check$BUGSoutput$summary #for ease in typing
dfcheck<-data.frame(Predicted=sumtab[meanrows,"mean"],
  Residual=sumtab[residrows,"mean"],
  SResid=sumtab[sresidrows,"mean"])
ggplot(dfcheck)+
  geom_point(aes(x=Predicted,y=Residual))+
  geom_abline(intercept=0,slope=0)+
  ggtitle("Residuals")
ggplot(dfcheck)+
  geom_point(aes(x=Predicted,y=SResid))+
  geom_abline(intercept=0,slope=0) +
  ylab("Standardized residual")+
  ggtitle("Standardized residuals")
ggplot(dfcheck,aes(sample=Residual))+
  geom_qq()+geom_qq_line()+
  ggtitle("QQNormal of Residuals")
ggplot(dfcheck,aes(x=SResid))+
  geom_histogram(binwidth=.2)+
  xlab("Standardized residuals")+
  ggtitle("Histogram of residuals")
```

## Show a sample of the data points the Chi-squared discrepancy measure. 

```{r}
dim(res1check$BUGSoutput$sims.matrix)
samplerow=sample(1:dim(res1check$BUGSoutput$sims.matrix)[1],400)
dfchi<-data.frame(Observed=res1check$BUGSoutput$sims.matrix[samplerow,
  "chi.square.obs"],
  Replicated=res1check$BUGSoutput$sims.matrix[samplerow,"chi.square.rep"])
ggplot(dfchi)+geom_point(aes(x=Observed,y=Replicated))+geom_abline(slope=1,intercept=0)+ggtitle("Chi squared discrepancy")
```

## Model checking numbers

```{r}
round(sumtab["p.value","mean"],3)   # Bayesian P value
range(sumtab[,"Rhat"])  # Find range of BGR statistics for parameters. Both should be close to 1.0
range(sumtab[,"n.eff"][sumtab[,"n.eff"]>1])  # Find range of effective sample sizes. Should be large (>400)
which(sumtab[,"Rhat"]>1.1)
```

## Null model to compare to linear model for model selection

```{r}
write("model{
  a ~ dnorm (0, 1.0E-6)  	# prior for the intercept
  prec ~ dgamma(0.001, 0.001)  # prior for the precision 
  for (i in 1:16)  # for each of the 16 lakes
  {
    ymean[i] <- a # the predicted y value for the null model
    CWD[i] ~ dnorm(ymean[i], prec) #  the actual y, normally distributed 
    LL[i]<- -log(1/sqrt(prec))-0.5*log(2*3.14159)-0.5*(CWD[i]-ymean[i])^2*prec
  }
}
",file= here("JAGS_mods", "box5.1null.txt"))
data1null<-list(CWD=cwd$CWD)
init1null<-list(list(a=0,  prec=100),list(a=1, prec=50))
res1null<-jags(data1null,init1null,
  parameters.to.save = c("a","prec","LL"),
  model.file = here("JAGS_mods", "box5.1null.txt"),
  n.chains=2,n.thin=10,n.iter=110000,n.burnin=10000)
```

## DIC, WAIC and LOOIC

```{r}
LLrows<-paste0("LL[",1:16,"]")
waic1<-waic(res1$BUGSoutput$sims.matrix[,LLrows])$estimates
waic1null<-waic(res1null$BUGSoutput$sims.matrix[,LLrows])$estimates
looic1<-loo(res1$BUGSoutput$sims.matrix[,LLrows])$estimates
looic1null<-loo(res1null$BUGSoutput$sims.matrix[,LLrows])$estimates

ICtable<-data.frame(Model=c("Regression","Null"),
                     DIC=c(res1$BUGSoutput$DIC,res1null$BUGSoutput$DIC),
                     pD=c(res1$BUGSoutput$pD,res1null$BUGSoutput$pD),
                     WAIC=c(waic1["waic","Estimate"],waic1null["waic","Estimate"]),
                     pWAIC=c(waic1["p_waic","Estimate"],waic1null["p_waic","Estimate"]),
                     LOOIC=c(looic1["looic","Estimate"],looic1null["looic","Estimate"]),
                     pLOOIC=c(looic1["p_loo","Estimate"],looic1null["p_loo","Estimate"]),
                     elpd=c(looic1["elpd_loo","Estimate"],looic1null["elpd_loo","Estimate"]))  %>% 
  mutate(deltaDIC=DIC-min(DIC),
    deltaWAIC=WAIC-min(WAIC),
    deltaLOOIC=LOOIC-min(LOOIC),
    WAIC.weight=exp(-0.5*deltaWAIC)/sum(exp(-0.5*deltaWAIC))) %>%
  mutate(across(where(is.numeric), round, 2))
ICtable
```

## STAN code for the same models

```{r}
standat1<-list(x=(cwd$TreeDens-mean(cwd$TreeDens))/sd(cwd$TreeDens),
  y=cwd$CWD,N=nrow(cwd))
stan1<-stan(file=here("STAN_docs","regression1.stan"),
  data=standat1
)
print(stan1)
stan_trace(stan1)
stan_dens(stan1)
```

## Stan with diagnostics

```{r}
stan2<-stan(file=here("STAN_docs", "regression2.stan"),
  data=standat1
)
print(stan2, pars=c("a","b","sigma","pvalue"))
```

## Get waic and looic

```{r}
ll1<-extract_log_lik(stan2,parameter_name = "loglik")
waic(ll1)
loo(ll1)
```

## Extract all values for chi-squared plot

```{r}
?extract
chi2vals<-extract(stan2,pars=c("chi2sim","chi2dat"))
chi2vals<-as.data.frame(chi2vals)
head(chi2vals)
ggplot(chi2vals,aes(x=chi2sim,y=chi2dat))+
  geom_point(alpha=0.4)+
  geom_abline(intercept=0,slope=1)
```

## get residuals and prediction for plots

```{r}
get_posterior_mean(stan2,pars="resid")
cwd$Residual<-get_posterior_mean(stan2,pars="resid")[,"mean-all chains"]
cwd$Predicted<-get_posterior_mean(stan2,pars="ypred")[,"mean-all chains"]
ggplot(cwd,aes(x=Predicted,y=Residual))+
  geom_point()+
  geom_hline(yintercept=0)
ggplot(cwd,aes(sample=Residual))+
  geom_qq()+geom_qq_line()
```

## Null and waic/looic comparison

```{r}
stan1null<-stan(file=here("STAN_docs", "regression1null.stan"),
  data=standat1[-1]
)
ll1null<-extract_log_lik(stan1null,parameter_name = "loglik")
waic(ll1null)
loo(ll1null)
```


