---
title: "Lecture 11. More on regression"
author: "Beth Babcock"
date: '`r Sys.Date()`'
output: html_document
---

```{r,message=FALSE}
library(tidyverse)
library(R2jags)
library(ggmcmc)
library(rstan)
library(loo)
library(gridExtra)
library(here)
options(mc.cores = parallel::detectCores())
# To keep a compiled version of the code so you don't have to recompile
rstan_options(auto_write = TRUE)
```

## Grasses example. Box 5.3 and 5.4

Set working directory to folder that includes data file 

```{r}
data5.3<-read.csv(here("data", "box5.3data.csv"))
data5.3$Y<-log(data5.3$C3+1)
summary(data5.3)
ggplot(data5.3,aes(x=LONG,y=LAT,size=C3))+
  geom_point(alpha=0.5)+
  ggtitle("Proportion of C3 grasses")+
  xlab("Longitude")+ylab("Latitude")
```


JAGS with no interaction

```{r}
write("model
{
  mLAT<-mean(LAT)
  mLONG<-mean(LONG)
  for (i in 1:73)			# for each of the 73 sites
  {
    Y[i] ~ dnorm(ymean[i], prec)	# assume normal distribution
    ymean[i] <- a + b[1]*(LONG[i]-mLONG) + b[2]*(LAT[i]-mLAT)
   pred.obs[i]~dnorm(ymean[i],prec)   # Predicted Y value
   resid[i]<-Y[i]-ymean[i]			# Residual 
   sresid[i]<-(Y[i]-ymean[i])*sqrt(prec)  # Standardized residual
   sresid2[i]<-sresid[i]*sresid[i]  # Pearson residual squared
   rep.sresid2[i]<-(pred.obs[i]-ymean[i])*(pred.obs[i]-ymean[i])*prec
   LL[i]<--0.5*log(2*3.14159)+0.5*log(prec)-0.5*prec*(Y[i]-ymean[i])*(Y[i]-ymean[i])
  }
  # uninformative priors
  a ~ dnorm(0, 1.0E-6)
  for (i in 1:2)
  {
    b[i] ~ dnorm(0, 1.0E-6)
  }
  prec ~ dgamma(0.001, 0.001)
  # make predictions over longitudes 93 to 120
  for (i in 93:120)
  {
    predlat35[i] <- a + b[1]*(i-mLONG) + b[2]*(35-mLAT)
    predlat45[i] <- a + b[1]*(i-mLONG) + b[2]*(45-mLAT)
    # back-transformed prediction at latitude 35
    predrichlat35[i] <- exp(predlat35[i]) - 1
    predrichlat45[i] <- exp(predlat45[i]) - 1
  }
 # Model checking
 sdy<-sd(Y[])
 sdy2<-sdy*sdy
 r.squared<-1-1/(prec*sdy2)
 chi.square.obs<-sum(sresid2[])
 chi.square.rep<-sum(rep.sresid2[])
 p.value<-step(chi.square.obs-chi.square.rep)
}
",file=here("JAGS_mods","box5.3.txt"))


res5.3<-jags(data5.3,
             parameters.to.save =  c("a","b","prec","predrichlat35","predrichlat45",
  "resid","ymean","sresid","p.value","r.squared","chi.square.obs","chi.square.rep","LL"),
  model.file = here("JAGS_mods","box5.3.txt"),
  n.chains=2,n.thin=10,n.iter=110000,n.burnin=10000)

round(res5.3$BUGSoutput$summary[c("a","b[1]","b[2]","p.value","r.squared"),],4)
```

## Box 5.4

Same with interaction

```{r}
write("model
{
  mLONG <- mean(LONG[])	# calculate mean of the two explanatory variables
  mLAT <- mean(LAT[])
  for (i in 1:73)			# for each of the 73 sites
  {
    Y[i] ~ dnorm(ymean[i], prec)	# assume normal distribution
    ymean[i] <- a + b[1]*(LONG[i]-mLONG) + b[2]*(LAT[i]-mLAT) + b[3]*(LONG[i]-mLONG)*(LAT[i]-mLAT)
    #model checking
    pred.obs[i]~dnorm(ymean[i],prec)   # Predicted Y value
    resid[i]<-Y[i]-ymean[i]			# Residual 
    sresid[i]<-(Y[i]-ymean[i])*sqrt(prec)  # Standardized residual
    sresid2[i]<-sresid[i]*sresid[i]  # Pearson residual squared
    rep.sresid2[i]<-(pred.obs[i]-ymean[i])*(pred.obs[i]-ymean[i])*prec
    LL[i]<--0.5*log(2*3.14159)+0.5*log(prec)-0.5*prec*(Y[i]-ymean[i])*(Y[i]-ymean[i])
 }
  # uninformative priors
  a ~ dnorm(0, 1.0E-6)
  for (i in 1:3)
  {
    b[i] ~ dnorm(0, 1.0E-6)
    bsig[i]<-step(b[i])
  }
  prec ~ dgamma(0.001, 0.001)

  # make predictions over longitudes 93 to 120
  for (i in 93:120)
  {
    predlat35[i] <- a + b[1]*(i-mLONG) + b[2]*(35-mLAT) + b[3]*(i-mLONG)*(35-mLAT)
    predlat45[i] <- a + b[1]*(i-mLONG) + b[2]*(45-mLAT) + b[3]*(i-mLONG)*(45-mLAT)

    # back-transformed prediction at latitude 35
    predrichlat35[i] <- exp(predlat35[i]) - 1
    predrichlat45[i] <- exp(predlat45[i]) - 1
  }
  # Model checking
  sdy<-sd(Y[])
  sdy2<-sdy*sdy
  r.squared<-1-1/(prec*sdy2)
  chi.square.obs<-sum(sresid2[])
  chi.square.rep<-sum(rep.sresid2[])
  p.value<-step(chi.square.obs-chi.square.rep)
}
",file=here("JAGS_mods","box5.4.txt"))

init1<-list(list(a=0, b=c(0, 0, 0), prec=100),list(a=1, b=c(1,1, 1), prec=30))
res5.4<-jags(data5.3,init1,parameters.to.save = c("a","b","prec","predrichlat35","predrichlat45",
  "resid","sresid","ymean","p.value","r.squared","LL","bsig"),
  model.file = here("JAGS_mods","box5.4.txt"),n.chains=2,n.thin=10,n.iter=110000,n.burnin=10000)

round(res5.4$BUGSoutput$summary[c("a","b[1]","b[2]","b[3]","p.value","r.squared","bsig[1]","bsig[2]","bsig[3]"),],3)
```

## Compare DIC, WAIC and LOO

```{r}
LL5.3<-res5.3$BUGSoutput$sims.matrix[,paste0("LL[",1:73,"]")]
waic5.3<-waic(LL5.3)$estimate
loo5.3<-loo(LL5.3)$estimate
LL5.4<-res5.4$BUGSoutput$sims.matrix[,paste0("LL[",1:73,"]")]
waic5.4<-waic(LL5.4)$estimate
loo5.4<-loo(LL5.4)$estimate

modtab<-data.frame(model=c(5.3,5.4),
  DIC=c(res5.3$BUGSoutput$DIC,res5.4$BUGSoutput$DIC),
  pD=c(res5.3$BUGSoutput$pD,res5.4$BUGSoutput$pD),
  WAIC=c(waic5.3[3,1],waic5.4[3,1]),
  pWAIC=c(waic5.3[2,1],waic5.4[2,1]),
  LOOIC=c(loo5.3[3,1],loo5.4[3,1]),
  pLOOIC=c(loo5.3[2,1],loo5.4[2,1])) %>%
  mutate(deltaDIC=DIC-min(DIC),
         deltaWAIC=WAIC-min(WAIC),
         deltaLOO=LOOIC-min(LOOIC))
round(modtab,2)

```

# Make plots using the values predicted at 35 and 45 latitude

```{r}
rowvals<-c(paste0("predrichlat35[",93:120,"]"),
 paste0("predrichlat45[",93:120,"]"))
length(rowvals)
restab5.4<-data.frame(res5.4$BUGSoutput$summary[rowvals,c("mean","2.5%","97.5%")])
restab5.4$Model=rep("Interaction",56)
restab5.4$Latitude=rep(c(35,45),each=28)
restab5.4$Longitude=rep(93:120,2)
restab5.3<-data.frame(res5.3$BUGSoutput$summary[rowvals,c("mean","2.5%","97.5%")])
restab5.3$Model=rep("No Interaction",56)
restab5.3$Latitude=rep(c(35,45),each=28)
restab5.3$Longitude=rep(93:120,2)
restab<-rbind(restab5.3,restab5.4)
names(restab)
restab$Latitude<-factor(restab$Latitude)
ggplot(restab)+geom_line(aes(x=Longitude,y=mean,group=Latitude,color=Latitude))+
  geom_ribbon(aes(x=Longitude,ymin=X2.5.,ymax=X97.5.,group=Latitude,fill=Latitude),alpha=0.4)+
  facet_wrap(Model~.)+ylab("Proportion C3")+theme_classic()
```

## Residual plots. 

Change n and sumtab to change model

```{r}
n<-length(data5.3$Y)
#sumtab<-res5.3$BUGSoutput$summary
sumtab<-res5.4$BUGSoutput$summary
residrows<-paste0("resid[",1:n,"]")
sresidrows<-paste0("sresid[",1:n,"]")
meanrows<-paste0("ymean[",1:n,"]")
dfcheck<-data.frame(Predicted=sumtab[meanrows,"mean"],
  Residual=sumtab[residrows,"mean"],
  SResid=sumtab[sresidrows,"mean"])
ggplot(dfcheck)+geom_point(aes(x=Predicted,y=Residual))+geom_abline(intercept=0,slope=0)+ggtitle("Residuals")
ggplot(dfcheck)+geom_point(aes(x=Predicted,y=SResid))+geom_abline(intercept=0,slope=0) +ylab("Standardized residual")+ggtitle("Standardized residuals")
ggplot(dfcheck,aes(sample=Residual))+geom_qq()+geom_qq_line()+ggtitle("QQNormal of Residuals")
ggplot(dfcheck,aes(x=SResid))+geom_histogram(binwidth=.2)+xlab("Standardize residuals")+ggtitle("Histogram of residuals")

```

## Normal and lognormal

```{r}
#Normal vs. lognormal likelihood
#1. normal on log transformed data
write("model
{
 # Normal model on log-transformed data log(Y)
 norm.mean ~ dnorm(0, 1.0E-6)  	
 norm.prec~dgamma(0.01,0.01) 
 norm.var<-1/norm.prec  
 for (i in 1:40)  					
  {
     logY[i]~dnorm(norm.mean, norm.prec)  
  }
 lnorm.mean<-exp(norm.mean+1/(2*norm.prec))  #Calculate the expected value and variance of the untranformed data
 lnorm.var<-exp(2*norm.mean)*exp(1/norm.prec)*(exp(1/norm.prec)-1)
 lnorm.median<-exp(norm.mean)  
}
",file=here("JAGS_mods", "normalLogY.txt"))

# Lognormal model on Y
write("model
{
norm.mean ~ dnorm(0, 1.0E-6)  	
norm.prec~dgamma(0.01,0.01)
norm.var<-1/norm.prec   
for (i in 1:40)  					
{
  Y[i]~dlnorm(norm.mean,norm.prec)
}
 lnorm.mean<-exp(norm.mean+1/(2*norm.prec))  #Calculate the expected value and variance of the untranformed data
 lnorm.var<-exp(2*norm.mean)*exp(1/norm.prec)*(exp(1/norm.prec)-1)
 lnorm.median<-exp(norm.mean)  
}
",file=here("JAGS_mods","lognormal.txt"))

#3, normal on Y
write("model
{
  # normal model on Y
  norm.mean ~ dnorm(0, 1.0E-6)  	
  norm.prec~dgamma(0.01,0.01)
  norm.var<-1/norm.prec   
  for (i in 1:40)  					
  {
    Y[i]~dnorm(norm.mean,norm.prec)
  }
}
",file=here("JAGS_mods","normal.txt"))


data1<-data.frame(Y=c(86.8, 182.1, 41.5, 66.4, 49.6, 109, 28.9, 56.3, 93.9, 80.2, 
                 30.2, 55.7, 57.6, 57.1, 49.1, 88, 22.9, 80.9, 144.5, 106, 69.2, 
                 19.2, 56.5, 34.3, 16.7, 129.9, 83.2, 64.8, 144.8, 73.3, 99, 49.7, 
                 33.5, 41.6, 32.6, 64.1, 23, 38.7, 89.7, 21.6))
data1$logY<-log(data1$Y)

res.normal.log<-jags(data1,parameters.to.save = c( "lnorm.mean",  "lnorm.var",  "lnorm.median","norm.mean","norm.var"),
                     model.file=here("JAGS_mods", "normalLogY.txt"),
                     n.burnin = 10000,n.chains = 2,n.iter = 210000,n.thin=10)            
round(res.normal.log$BUGSoutput$summary,2)

res.lognormal<-jags(data1,parameters.to.save = c( "lnorm.mean",  "lnorm.var",  "lnorm.median","norm.mean","norm.var"),
                     model.file=here("JAGS_mods","lognormal.txt"),
                     n.burnin = 10000,n.chains = 2,n.iter = 210000,n.thin=10)            
round(res.lognormal$BUGSoutput$summary,2)

res.normal<-jags(data1,parameters.to.save = c("norm.mean","norm.var"),
                    model.file=here("JAGS_mods","normal.txt"),
                    n.burnin = 10000,n.chains = 2,n.iter = 210000,n.thin=10)            
round(res.normal$BUGSoutput$summary,2)

dictab<-data.frame(model=c("normal on log","lognormal","normal"),
                   DIC=c(res.normal.log$BUGSoutput$DIC,res.lognormal$BUGSoutput$DIC,
                         res.normal$BUGSoutput$DIC))
dictab

```


## C4 grasses not log transformed

```{r}
data5.3norm<-data5.3 %>% mutate(Y=C3)
res5.3norm<-jags(data5.3norm,
             parameters.to.save =  c("a","b","prec","predrichlat35","predrichlat45",
  "resid","ymean","sresid","p.value","r.squared","chi.square.obs","chi.square.rep","LL"),
  model.file = here("JAGS_mods","box5.3.txt"),
  n.chains=2,n.thin=10,n.iter=110000,n.burnin=10000)

n<-length(data5.3$Y)
sumtab<-res5.3norm$BUGSoutput$summary
residrows<-paste0("resid[",1:n,"]")
meanrows<-paste0("ymean[",1:n,"]")
dfcheck<-data.frame(Predicted=sumtab[meanrows,"mean"],
  Residual=sumtab[residrows,"mean"])
g1<-ggplot(dfcheck)+geom_point(aes(x=Predicted,y=Residual))+
  geom_abline(intercept=0,slope=0)+ggtitle("Residuals")
g2<-ggplot(dfcheck,aes(sample=Residual))+
  geom_qq()+geom_qq_line()+
  ggtitle("QQNormal of Residuals")
grid.arrange(g1,g2,ncol=2)

waic(res5.3norm$BUGSoutput$sims.matrix[,paste0("LL[",1:73,"]")])$estimate[3,1]-
waic(res5.3$BUGSoutput$sims.matrix[,paste0("LL[",1:73,"]")])$estimate[3,1]
```
## To compare non-linear to linear regression

 Box 5.6  Non-linear regression of CWD 
 
```{r}
cwd<-data.frame(TreeDens=c(1270, 1210, 1800, 1875, 1300, 2150, 1330, 964, 961, 1400, 1280, 976, 771, 833, 883, 956),
  CWD=c(121, 41, 183, 130, 127, 134, 65, 52, 12, 46, 54, 97, 1, 4, 1, 4))

write("model{
  a ~ dnorm (0, 1.0E-6)  	# prior for the intercept
  b ~ dnorm (0, 1.0E-6)  		# prior for the slope
  prec ~ dgamma(0.001, 0.001)  # prior for the precision 
  for (i in 1:16)  # for each of the 16 lakes
  {
    ymean[i] <- a + b*TreeDens[i]  # the predicted y value y=a+bx
    CWD[i] ~ dnorm(ymean[i], prec) #  the actual y, normally distributed 
    pred.obs[i] ~ dnorm(ymean[i], prec) # with error added
    resid[i]<-CWD[i]-ymean[i]	
    sresid[i]<-(CWD[i]-ymean[i])*sqrt(prec) 
    sresid2[i]<-sresid[i]*sresid[i] 
    rep.sresid2[i]<-(pred.obs[i]-ymean[i])*(pred.obs[i]-ymean[i])*prec
    LL[i]<- -log(1/sqrt(prec))-0.5*log(2*3.14159)-0.5*(CWD[i]-ymean[i])^2*prec
  }
  #summary statistics
  prob.pos.b<-step(b)
  sdy<-sd(CWD[])
  sdy2<-sdy*sdy
  r.squared<-1-1/(prec*sdy2)
  chi.square.obs<-sum(sresid2[])
  chi.square.rep<-sum(rep.sresid2[])
  p.value<-step(chi.square.obs-chi.square.rep)
}
",file=here("JAGS_mods","box5.1.txt"))

data1<-list(TreeDens=cwd$TreeDens,
  CWD=cwd$CWD,
  x.new=c(800,1000,1200,1400,1600,1800,2000,2200), #New data for predicting
  n.new=8)

#fit model
res1<-jags(data1,
  parameters.to.save = c("a","b","prec",
  "ymean","pred.obs","prob.pos.b","r.squared","LL"),
  model.file = here("JAGS_mods","box5.1.txt"),
  n.chains=2,n.thin=10,n.iter=110000,n.burnin=10000)
res1sum<-data.frame(res1$BUGSoutput$summary) 
res1sum$parameter=rownames(res1sum)
res1sum<-filter(res1sum,grepl("ymean",parameter))
res1sum<-bind_cols(res1sum,cwd)

#nonlinear
write("model {
    a ~ dunif (0, 1000)					# uninformative prior, constrained to be positive
    b ~ dnorm (0, 1.0E-6)				# uninformative prior
    c ~ dunif(0, 10)							# uninformative prior, constrained to be positive
    prec ~ dgamma(0.001, 0.001)	# uninformative prior
    mTree <- mean(TreeDens[])  	# mean of the explanatory variable
    
    for (i in 1:16)  # for each of the 16 lakes
    {
      predm[i] <- a / (1 + exp(-(b+c*(TreeDens[i]-mTree))))  # predicted CWD
      logpred[i] <- log(predm[i])  # take the logarithm of the prediction
      CWD[i] ~ dlnorm(logpred[i], prec)  # CWD drawn from lognormal
      logCWD[i]<-log(CWD[i])
      pred.obs[i]~dlnorm(logpred[i],prec)   # Predicted CWD at this tree density
      resid[i]<-logCWD[i]-logpred[i]			# Residual CWD at this tree density
      sresid[i]<-(logCWD[i]-logpred[i])*sqrt(prec)  # Standardized residual
      sresid2[i]<-sresid[i]*sresid[i]  # Pearson residual squared
      rep.sresid2[i]<-(log(pred.obs[i])-logpred[i])*(log(pred.obs[i])-logpred[i])*prec
    }
    prob.pos.b<-step(b)
    chi.square.obs<-sum(sresid2[])
    chi.square.rep<-sum(rep.sresid2[])
    p.value<-step(chi.square.obs-chi.square.rep)
    # make predictions at new values for plotting
    for(j in 1:n.new) {
      mean.new[j] <- a / (1 + exp(-(b+c*(x.new[j]-mTree))))  # predicted CWD
      logpred.new[j] <- log(mean.new[j])  # take the logarithm of the prediction
      pred.new[j] ~ dlnorm(logpred.new[j], prec)  # CWD drawn from lognormal
     }
  }",here("JAGS_mods","box5.6.txt"))

init1=list(list(a=2, b=0, c=1, prec=10),list(a=1.5, b=0.5, c=0.5, prec=5))
params=c("a","b","c","prec","p.value","resid","predm","pred.obs","sresid","prob.pos.b","chi.square.obs","chi.square.rep")
res5.6=jags(data1,init1,model.file=here("JAGS_mods","box5.6.txt"),
  parameters.to.save=params,n.chains=2,n.iter=510000,n.burnin=10000,n.thin=20)

range(res5.6$BUGSoutput$summary[,"Rhat"])  # Find range of BGR statistics for parameters. Both should be close to 1.0
range(res5.6$BUGSoutput$summary[,"n.eff"])  # Find range of effective sample sizes. SHould be large (>400)
res5.6$BUGSoutput$summary[c("a","b","c","p.value"),]

# To make model checking and diagnostic plots
n<-length(data1$TreeDens)
sumtab<-res5.6$BUGSoutput$summary
residrows<-paste0("resid[",1:n,"]")
meanrows<-paste0("predm[",1:n,"]")
dfcheck<-data.frame(Predicted=sumtab[meanrows,"mean"],
  Residual=sumtab[residrows,"mean"])
ggplot(dfcheck)+geom_point(aes(x=Predicted,y=Residual))+geom_abline(intercept=0,slope=0)+ggtitle("Residuals")
ggplot(dfcheck,aes(sample=Residual))+geom_qq()+geom_qq_line()+ggtitle("QQNormal of Residuals")
#Compare DIC to linear model
res5.6$BUGSoutput$DIC
res5.6$BUGSoutput$pD
dictab<-data.frame(Model=c("Logistic","Linear"),
  DIC=c(res5.6$BUGSoutput$DIC,res1$BUGSoutput$DIC),
  pD=c(res5.6$BUGSoutput$pD,res1$BUGSoutput$pD))
dictab$deltaDIC=dictab$DIC-min(dictab$DIC)
dictab

#Plot both fits
dflog<-data.frame(TreeDens=data1$TreeDens,CWD=data1$CWD)

dflog$Predict<-res5.6$BUGSoutput$summary[meanrows,"mean"]
dflog$lci<-res5.6$BUGSoutput$summary[meanrows,"2.5%"]
dflog$uci<-res5.6$BUGSoutput$summary[meanrows,"97.5%"]

ggplot(dflog,aes(x=TreeDens))+geom_point(aes(y=CWD))+
  geom_line(aes(y=Predict),col="red")+
  geom_ribbon(aes(x=TreeDens,ymin=lci,ymax=uci),alpha=0.3,fill="red")+
  xlab("Tree density")+ylab("Course woody debris")+
  geom_line(data=res1sum,aes(x=TreeDens,y=X50.),col="blue")+
  geom_ribbon(data=res1sum,aes(x=TreeDens,ymin=X2.5.,ymax=X97.5.),fill="blue",alpha=0.3)+theme_bw()

```
## Matrix format

```{r}
grass<-data5.3%>%
  mutate(LONG=(LONG-mean(LONG))/sd(LONG),LAT=(LAT-mean(LAT))/sd(LAT))
mod1<-lm(Y~LONG*LAT,data=grass)
grassMatrix<-model.matrix(mod1)
head(grassMatrix)

write("model{
  for (i in 1:Ncoef)  {
    b[i] ~ dnorm(0, 1.0E-6)
  }
  prec ~ dgamma(0.001, 0.001)
  for (i in 1:N)		
  {
    ymean[i] <- inprod(b,xMatrix[i,])
    Y[i] ~ 	dnorm(ymean[i],prec)
  }
}
",file=here("JAGS_mods","box5.3matrix.txt"))
  
grassData<-list(Y=grass$Y,xMatrix=grassMatrix,Ncoef=4,N=nrow(grass))
mod1<-jags(data=grassData,
  parameters.to.save=c("b","prec"),
  model.file=here("JAGS_mods","box5.3matrix.txt"),
  n.chain=2,n.iter=110000,n.burnin=10000,n.thin=4)
print(mod1)
```

## Same in STAN

```{r}
write("
data{
 int N;
 int Ncoef;
 vector[N] Y;
 matrix[N,Ncoef] xMatrix;
}
parameters{
 vector[Ncoef] b;
 real<lower=0> Sigma;
}
model{
  b~normal(0,10);
  Sigma~exponential(1);
  Y~normal(xMatrix*b,Sigma);
}
",file=here("STAN_docs", "box5.3matrix.stan"))
box5.3stan<-stan(file=here("STAN_docs", "box5.3matrix.stan"),
  data=grassData)
print(box5.3stan)
```