---
title: "12. GLM"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

## Lecture 12. GLM

```{r,message=FALSE}
library(tidyverse)
library(R2jags)
library(ggmcmc)
library(rstan)
library(loo)
library(gridExtra)
library(DHARMa)
library(here)
theme_set(theme_bw())
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

Plots for GLM mean to variance

```{r,warnings=FALSE}
# Beta conversions
betab<-function(mu,a) a*(1-mu)/mu
betase<-function(a,b) sqrt(a*b/((a+b)^2*(a+b+1)))

#Plot of residual variance vs. mean 
plotdat<-expand.grid(Mean=seq(0,1,by=0.02),
  Distribution=c("Normal","Poisson","Negative Binomial","Beta"),Variance=0.5) %>%
  mutate(Variance=ifelse(Distribution=="Poisson",Mean,Variance),
         Variance=ifelse(Distribution=="Negative Binomial",Mean+Mean^2/3,Variance),
         Variance=ifelse(Distribution=="Beta",betase(2,betab(Mean,2)),Variance))
ggplot(plotdat,aes(x=Mean,y=Variance,color=Distribution))+
  geom_line(lwd=2)
```

## Box 5.10 Poisson with log link

```{r}
ant<-read.csv(here("data","ant.csv"))
antlong<-ant %>% pivot_longer(c("lat","elev","habitat"),values_to="Value",names_to="Variable")
ggplot(antlong,aes(x=Value,y=richness))+
  geom_smooth(method=lm)+
  geom_point()+
  facet_wrap(Variable~.,scales="free")
ggplot(antlong,aes(x=Value,y=richness))+
  geom_smooth(method=glm,method.args = list(family = "poisson"))+
  geom_point()+
  facet_wrap(Variable~.,scales="free")
#?glm
```

```{r}
write("model 
{ 
# uses centred data

  ml <- mean(lat[])  # calculates averages for centring
  me <- mean(elev[])

  for( i in 1 : N)  # for each of the N data points
  { 
    richness[i] ~ dpois(mu[i])  # ant species richness drawn from a Poisson
    log(mu[i]) <- alpha + beta[1]*(lat[i]-ml) + beta[2]*(elev[i]-me) + beta[3]*habitat[i]  # a log-linear model for the average
    lresid[i]<-log(richness[i])-log(mu[i]) #Works if no zero values in richness    
    resid[i]<-richness[i]-mu[i]  
    sresid[i]<-(richness[i]-mu[i] )/sqrt(mu[i]) #Pearson residual
    simRichness[i]~dpois(mu[i])
  } 
  intercept <- alpha - beta[1]*ml - beta[2]*me  # recovers the intercept term
# informative priors
  preclat <- 1 / (0.04 * 0.04)  # precision of lat effect, based on s.e. of 0.04
  precelev <- 1 / (0.0003 * 0.0003)  # precision of elev effect
  alpha ~ dnorm(0.0,1.0E-6)  # uninformative intercept
  beta[1] ~ dnorm(-0.1725,preclat)  # informative effect of latitude
  beta[2] ~ dnorm(-0.0022,precelev)  # informative effect of elevation
  beta[3] ~ dnorm(0.0,1.0E-6)  # uninformative effect of habitat
}",file=here("JAGS_mods","box5.10.txt"))
data5.10<-c(ant,N=nrow(ant))
init1<-list(list(alpha=0, beta=c(0,0,0)),list(alpha=.1, beta=c(.1,.1,.1)))
res5.10<-jags(data5.10,init1,
  model.file=here("JAGS_mods","box5.10.txt"),
  parameters.to.save=c(c("alpha","beta","mu","intercept","lresid","resid","sresid","simRichness")),
  n.chains=2,n.iter=310000,n.burnin=10000,n.thin=20)

range(res5.10$BUGSoutput$summary[,"Rhat"])  
range(res5.10$BUGSoutput$summary[,"n.eff"]) 
res5.10$BUGSoutput$summary[c("alpha","beta[1]","beta[2]","beta[3]","intercept"),]
```

## Plot residuals

```{r}
ant$mu<-res5.10$BUGSoutput$summary[paste0("mu[",1:data5.10$N,"]"),"mean"]
ant$lresid<-res5.10$BUGSoutput$summary[paste0("lresid[",1:data5.10$N,"]"),"mean"]
ant$resid<-res5.10$BUGSoutput$summary[paste0("resid[",1:data5.10$N,"]"),"mean"]
ant$sresid<-res5.10$BUGSoutput$summary[paste0("sresid[",1:data5.10$N,"]"),"mean"]
g1<-ggplot(ant)+
  geom_point(aes(x=log(mu),y=lresid))+
  xlab("Predicted log mean")+ylab("Residuals in log scale")+
  geom_abline(aes(intercept=0,slope=0))+
  ggtitle("Log scale residual")
g2<-ggplot(ant)+
  geom_point(aes(x=mu,y=resid))+
  xlab("Predicted mean")+
  ylab("Residuals in original scale")+
  geom_abline(aes(intercept=0,slope=0))+ggtitle("Not log")
g3<-ggplot(ant)+geom_point(aes(x=mu,y=resid))+
  xlab("Predicted mean")+
  ylab("Pearson residuals")+
  geom_abline(aes(intercept=0,slope=0))+
  ggtitle("Pearson residual")
grid.arrange(g1,g2,g3,ncol=3)
```

## DHARMa residuals. 

See here https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html for details on how to implement. DHARMa does this automatically for glm, glmer, gam, glmmTMB, etc. 

```{r}
mod1<-glm(richness~lat+elev+habitat,data=ant,family="poisson")
plot(simulateResiduals(mod1))
```

For Bayesian models:

1. pull out a matrix of predicted values from the MCMC draws. and transpose so data points are rows and draws are columns
2. Sample the MCMC draws so you only have 1000, or the plot will be too busy
3. Create DHARMa object
4. Make plot

```{r}
#Get MCMC draws of the simulated richness
simRichness<-t(res5.10$BUGSoutput$sims.matrix[,paste0("simRichness[",1:data5.10$N,"]")])
dim(simRichness)
#Get a smaller sample
simRichness<-simRichness[,sample(1:30000,1000)]
#Pull out predicted mean values from summary
predRichness<-res5.10$BUGSoutput$summary[paste0("mu[",1:data5.10$N,"]"),"mean"]
#?createDHARMa
DHARMaRes <- createDHARMa(simulatedResponse =simRichness , 
                  observedResponse = data5.10$richness, 
             fittedPredictedResponse = predRichness, 
  integerResponse = TRUE)
#Make DHARMa plot
plot(DHARMaRes,quantreg=FALSE)
```

## Logit example plot

```{r}
logit<-function(p) log(p/(1-p))
ilogit<-function(x) exp(x)/(1+exp(x))
plotdat<-data.frame(logit.p=rnorm(100)) %>%
  mutate(p=ilogit(logit.p))
ggplot(plotdat,aes(x=logit.p,y=p))+geom_line(lwd=2)
```

## Box 5.5 Bird mortality. Using logit transformation for proportion data  

```{r}
bird<-read.csv(here("data","birdMortality.csv"))
summary(bird)
bird$Owl<-factor(bird$Owl)
ggplot(bird)+geom_point(aes(x=Mass,y=Mortality,col=Owl))+theme_bw()
```

## Model in JAGS

```{r}
write("model {
for (i in 1:26)  # for each of the 26 raptor species
{
  lp[i] <- a + b*Mass[i] + c*Owl[i]  # linear predictor
  lm[i] ~ dnorm(lp[i], prec)  # assume logit(Mortality) is normally distributed
  pred.mort[i]<-exp(lp[i])/(1+exp(lp[i]))
  pred.obs[i]~dnorm(lp[i],prec)   # Predicted Y value
  resid[i]<-lm[i]-lp[i]			# Residual 
  sresid[i]<-(lm[i]-lp[i])*sqrt(prec)  # Standardized residual
  sresid2[i]<-sresid[i]*sresid[i]  # Pearson residual squared
  rep.sresid2[i]<-(pred.obs[i]-lp[i])*(pred.obs[i]-lp[i])*prec
}
a ~ dnorm(0, 1.0E-6)  # intercept term
b ~ dnorm(0, 1.0E-6)  # effect of body mass
c ~ dnorm(0, 1.0E-6)  # effect of being an owl
prec ~ dgamma(0.001, 0.001)  # precision
chi.square.obs<-sum(sresid2[])
chi.square.rep<-sum(rep.sresid2[])
p.value<-step(chi.square.obs-chi.square.rep)
}",here("JAGS_mods","Model5.5.txt"))

init1=list(a=0, b=0, c=0, prec=20)
init2=list(a=1, b=1, c=1, prec=50)

bird$lm<-log(bird$Mortality/(1-bird$Mortality)) # Make logit of mortality for JAGS
  
res5.5<-jags(bird[,c("Mass","Owl","lm")],
             list(init1,init2),
             model.file=here("JAGS_mods","Model5.5.txt"),
            parameters.to.save=c("a","b","c","prec","p.value","resid","pred.mort","lp","pred.obs","sresid","chi.square.obs","chi.square.rep"),
  n.chains=2,n.iter=110000,n.burnin=10000,n.thin=20)

range(res5.5$BUGSoutput$summary[,"Rhat"])  # Find range of BGR statistics for parameters. Both should be close to 1.0
range(res5.5$BUGSoutput$summary[,"n.eff"])  # Find range of effective sample sizes. SHould be large (>400)
```

Plot residuals

```{r}
n<-dim(bird)[1]
sumtab<-res5.5$BUGSoutput$summary
residrows<-paste0("resid[",1:n,"]")
meanrows<-paste0("pred.obs[",1:n,"]")
dfcheck<-data.frame(Predicted=sumtab[meanrows,"mean"],
  Residual=sumtab[residrows,"mean"])
g1<-ggplot(dfcheck)+
  geom_point(aes(x=Predicted,y=Residual))+
  geom_abline(intercept=0,slope=0)+ggtitle("Residuals")
g2<-ggplot(dfcheck,aes(sample=Residual))+
  geom_qq()+geom_qq_line()+ggtitle("QQNormal of Residuals")
grid.arrange(g1,g2,ncol=2)
res5.5$BUGSoutput$summary["p.value","mean"]
```

Plot fit in probability space

```{r}
predrows<-paste0("pred.mort[",1:n,"]")
bird$Predict<-res5.5$BUGSoutput$summary[predrows,"mean"]
bird$lci<-res5.5$BUGSoutput$summary[predrows,"2.5%"]
bird$uci<-res5.5$BUGSoutput$summary[predrows,"97.5%"]
ggplot(bird)+geom_point(aes(x=Mass,y=Mortality,col=Owl))+theme_bw()+
 geom_line(aes(x=Mass,y=Predict,col=Owl)) +
 geom_ribbon(aes(x=Mass,ymin=lci,ymax=uci,fill=Owl),alpha=0.3)
round(res5.5$BUGSoutput$summary[c("a","b","c"),c("mean","sd","2.5%","50%","97.5%")],3)
```

## Box 5.7 logistic regression

```{r}
write("model
{
  a ~ dnorm(0, 1.0E-6)  # the intercept term
  for (i in 1:6)  # the 6 regression coefficients
  {
    b[i] ~ dnorm(0, 1.0E-6)  # uninformative priors
  }

  mnr <- mean(numrock[])  # means of the explanatory variables
  mv <- mean(volcanic[])
  mrc <- mean(rockcell[])
  msl <- mean(nearslope[])
  mso <- mean(nearsouth[])
  for (i in 1:325)  # for each of the sites
  {
    logit(p[i]) <- a + b[1]*(numrock[i]-mnr) + b[2]*(volcanic[i]-mv) + b[3]*(rockcell[i]-mrc) + b[4]*(nearslope[i]-msl) + b[5]*(nearslope[i]-msl)*(nearslope[i]-msl) + b[6]*(nearsouth[i]-mso)
  # logit(p) is a function of the explanatory variables and regression coeff's
    lr[i] ~ dbern(p[i])  # observed occurrence drawn from a Bernoulli dist'n
    resid[i]<-lr[i]-p[i]
    simval[i]~dbern(p[i])
  }
}",file=here("JAGS_mods","box5.7.txt"))

init1<-list(a=0, b=c(0,0,0,0,0,0))
init2<-list(a=0.1, b=c(0.1,0.1,0.1,0.1,0.1,0.1))

lr<-read.csv(here("data","lr.csv"))

res5.7<-jags(lr, list(init1,init1), 
               parameters.to.save=c("p","a","b","resid","simval"), 
               model.file=here("JAGS_mods","box5.7.txt"), n.chains=2, 
               n.iter=110000, n.burnin=10000,
               n.thin=2)

#Check convergence
range(res5.7$BUGSoutput$summary[,"Rhat"])
range(res5.7$BUGSoutput$summary[,"n.eff"])
```

# Coefficients

```{r}
mc1<-ggs(as.mcmc(res5.7))
ggs_caterpillar(filter(mc1,grepl("b",Parameter)),sort=FALSE)
```

Pull out predictions in probability space

```{r}
n<-dim(lr)[1]
prows<-paste0("p[",1:n,"]")   #make list of names of probabilities
lr$p<-res5.7$BUGSoutput$summary[prows,"mean"]
round(res5.7$BUGSoutput$summary[prows[1:4],],3)
ggplot(lr,aes(x=p,y=lr))+
  geom_point(alpha=0.3)
ggplot(lr,aes(x=factor(lr),y=p))+
  geom_boxplot(fill="grey")+
  xlab("Presence of lr")+
  ylab("Predicted presence")

```

Confusion matrix

```{r}
confusion<-table(round(lr$p),lr$lr)
confusion
(confusion[1,1]+confusion[2,2])/sum(confusion)
```

Residuals

```{r}
residrows<-paste0("resid[",1:n,"]")
lr$Residual<-res5.7$BUGSoutput$summary[residrows,"mean"]
ggplot(lr,aes(x=p,y=Residual))+geom_point()

#DHARMa residuals. 
simval<-t(res5.7$BUGSoutput$sims.matrix[,paste0("simval[",1:nrow(lr),"]")])
dim(simval)
simval<-simval[,sample(1:80000,500)]
DHARMaRes = createDHARMa(simulatedResponse =simval , observedResponse = lr$lr, 
             fittedPredictedResponse = lr$p, integerResponse = T)
plot(DHARMaRes)
```

