---
title: "13. ANOVA"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

```{r,message=FALSE}
library(R2jags)
library(tidyverse)
library(ggmcmc)
library(rstan)
library(loo)
library(here)
theme_set(theme_bw())
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```


## Starling anova box 6.1

```{r}
starling<-read.csv(here("data","starling6.1.csv"))
summary(starling)

ggplot(starling)+
  geom_boxplot(aes(group=factor(location),x=location,y=Y))+ylab("Starling mass")


```
```{r}
write("model   
{  # simple one-way ANOVA
  for (i in 1:4)  # the mean for each of the four locations, drawn from uninformative priors
  {
    a[i] ~ dnorm(0.0, 1.0E-6)
  }
  tau ~ dgamma(0.001, 0.001)  # uninformative precision
  for (i in 1:40) # for each of the 40 birds sampled
  {
    mean[i] <- a[location[i]]  # the mean depends on the location
    Y[i] ~ dnorm(mean[i], tau)  # mass drawn from a normal distribution
    Yrep[i]~ dnorm(mean[i], tau)
    resid[i]<-Y[i]-mean[i]
    sresid2[i]<-resid[i]*resid[i]*tau
    sresid2rep[i]<-(Yrep[i]-mean[i])*(Yrep[i]-mean[i])*tau
  }
    p.value<-step(sum(sresid2rep)-sum(sresid2))
}
",file=here("JAGS_mods","box6.1.txt"))
res6.1<-jags(starling,
             model.file=here("JAGS_mods","box6.1.txt"),
             parameters.to.save=c("a","tau","resid","mean","p.value"),
             n.chains=2,n.iter=110000,n.burnin=10000,n.thin=10)
round(res6.1$BUGSoutput$summary[c("a[1]","a[2]","a[3]","a[4]","tau","p.value"),],2)
gg1<-ggs(as.mcmc(res6.1))
gg1<-filter(gg1,Parameter%in%c("a[1]","a[2]","a[3]","a[4]"))
ggs_caterpillar(gg1)
#plot residuals
n<-nrow(starling)
sumtab<-res6.1$BUGSoutput$summary
residrows<-paste0("resid[",1:n,"]")
meanrows<-paste0("mean[",1:n,"]")
dfcheck<-data.frame(Predicted=sumtab[meanrows,"mean"],
  Residual=sumtab[residrows,"mean"])
ggplot(dfcheck)+geom_point(aes(x=Predicted,y=Residual))+geom_abline(intercept=0,slope=0)+ggtitle("Residuals")
ggplot(dfcheck,aes(sample=Residual))+geom_qq()+geom_qq_line()+ggtitle("QQNormal of Residuals")
```

## Box 6.1 Null for DIC comparison

```{r}
write("model   {     
  a ~ dnorm(0.0, 1.0E-6)
  tau ~ dgamma(0.001, 0.001)  # uninformative precision
  for (i in 1:40) # for each of the 40 birds sampled
  {
    mean[i] <- a
    Y[i] ~ dnorm(mean[i], tau)  # mass drawn from a normal distribution
  }
}
",file=here("JAGS_mods","box6.1null.txt"))
res6.1null=jags(starling,
                model.file=here("JAGS_mods","box6.1null.txt"),
                parameters.to.save=c("a","tau"),
                n.chains=2,n.iter=110000,n.burnin=10000,n.thin=10)
res6.1$BUGSoutput$DIC
res6.1$BUGSoutput$pD
res6.1null$BUGSoutput$DIC
res6.1null$BUGSoutput$pD
dictab<-data.frame(Model=c("anova","null"),
  DIC=round(c(res6.1$BUGSoutput$DIC,res6.1null$BUGSoutput$DIC),2),
  pD=round(c(res6.1$BUGSoutput$pD,res6.1null$BUGSoutput$pD),2))
dictab$deltaDIC<-dictab$DIC-min(dictab$DIC)
dictab
```


## Treatment contrasts in R for comparison

```{r}
starling$location<-factor(starling$location)
starlinglm<-lm(Y~location,data=starling)
summary(starlinglm)
starlinglm2<-lm(Y~location-1,data=starling)
summary(starlinglm2)
starling%>% group_by(location) %>% summarize(mean=mean(Y))
#intercept/base is the first level
starlingMatrix<-model.matrix(starlinglm)
starlingMatrix

```

## Box 6.2 with treatment contrast

```{r}
write("model  {  # one-way ANOVA, using reference class
  base ~ dnorm(0, 1.0E-6)  # the mean of the reference class
  for (i in 1:3)  # the mean for first 3 locations, relative to the reference class
  {
    d[i] ~ dnorm(0.0, 1.0E-6)
  }
  d[4] <- 0  # no diff b/t the mean of the 4th location and the reference class
  tau ~ dgamma(0.001, 0.001)  # uninformative precision
  for (i in 1:40) # for each of the 40 bids sampled
  {
    mean[i] <- base + d[location[i]]  # the mean depends on the location
    Y[i] ~ dnorm(mean[i], tau)
  }
}
",file=here("JAGS_mods","box6.2.txt"))

res6.2<-jags(starling,
             model.file=here("JAGS_mods","box6.2.txt"),
             parameters.to.save=c("base","d","tau"),
             n.chains=2,n.iter=110000,n.burnin=10000,n.thin=5)
round(res6.2$BUGSoutput$summary[c("base","d[1]","d[2]","d[3]","d[4]","tau"),],2)
res6.2$BUGSoutput$DIC
res6.2$BUGSoutput$pD
gg2<-ggs(as.mcmc(res6.2))
gg2<-filter(gg2,Parameter%in%c("base","d[1]","d[2]","d[3]"))
ggs_caterpillar(gg2)
ggs_pairs(gg2, lower=list(continuous="density"))
```

# Box 6.2 overparameterized to show why this is wrong. 

```{r}
write("model  {  # one-way ANOVA, with reference class wrong
  base ~ dnorm(0, 1.0E-6)  # the mean of the reference class
  for (i in 1:4)  # the mean each location, relative to the reference class
  {
    d[i] ~ dnorm(0.0, 1.0E-6)
  }
  tau ~ dgamma(0.001, 0.001)  # uninformative precision
  for (i in 1:40) # for each of the 40 bids sampled
  {
    mean[i] <- base + d[location[i]]  # the mean depends on the location
    Y[i] ~ dnorm(mean[i], tau)
  }
}
",file=here("JAGS_mods","box6.2wrong.txt"))
res6.2wrong=jags(starling,
                 model.file=here("JAGS_mods","box6.2wrong.txt"),
                 parameters.to.save=c("base","d","tau"),
                 n.chains=2,n.iter=110000,n.burnin=10000,n.thin=5)
round(res6.2wrong$BUGSoutput$summary[c("base","d[1]","d[2]","d[3]","d[4]","tau"),],2)
gg2wrong<-ggs(as.mcmc(res6.2wrong))
ggs_pairs(filter(gg2wrong,Parameter %in% c("base","d[1]","d[2]","d[3]","d[4]")))
```

## Contrasts in multi way ANOVA.

```{r}
starling2<-read.csv(here("data","starling.csv"))
starling2$period<-factor(starling2$period)
starling2$location<-factor(starling2$location)
ggplot(starling2,aes(x=location,y=mass,color=factor(period),group=period))+
  stat_summary()
starlinglm2<-lm(mass~period*location,data=starling2)
summary(starlinglm2)
view(model.matrix(starlinglm2))
```

## Starling with interactions in JAGS

```{r}
write("model   
{  # two-way ANOVA with interaction term, using reference classes
  base ~ dnorm(0, 1.0E-6)
  for (i in 1:3)  # the mean for each of the four locations, expressed as a difference from the reference class
  {
    a[i] ~ dnorm(0.0, 1.0E-6)
  }
  a[4] <- 0
# the mean for each of the two time periods, expressed as a difference from the reference class
# the first time period is used as the reference class
  b[1] <- 0
  b[2] ~ dnorm(0.0, 1.0E-6)

# interaction terms - these are set to zero for cases involving one or more reference classes
  for (i in 1:3)  
  {
    int[i,1] <- 0
    int[i, 2] ~ dnorm(0.0, 1.0E-6)
  }
  int[4,1] <- 0
  int[4,2] <- 0

  tau ~ dgamma(0.001, 0.001)  # uninformative precision

  for (i in 1:80) # for each of the 80 birds sampled
  {
    mean[i] <- base + a[location[i]] + b[period[i]] + int[location[i], period[i]] # the mean depends on the location and period
    Y[i] ~ dnorm(mean[i], tau)
  }
}
",file=here("JAGS_mods","box6.5.txt"))

## Interactions initial values for 2 way anova
starling2$Y<-starling2$mass
res6.5<-jags(starling2[,c("Y","location","period")],
             model.file=here("JAGS_mods","box6.5.txt"),
  parameters.to.save=c("base","a","b","int","tau"),
  n.chains=2,n.iter=110000,n.burnin=10000,n.thin=10)
round(res6.5$BUGSoutput$summary,2)
```

## Starling model without interactions

```{r}
write("model   
{  # two-way ANOVA with no interaction term, using reference classes
  
  base ~ dnorm(0, 1.0E-6)
  for (i in 1:3)  # the mean for each of the four locations, expressed as a difference from the reference class
  {
    a[i] ~ dnorm(0.0, 1.0E-6)
  }
  a[4] <- 0
# the mean for each of the two time periods, expressed as a difference from the reference class
# the first time period is used as the reference class
  b[1] <- 0
  b[2] ~ dnorm(0.0, 1.0E-6)
  tau ~ dgamma(0.001, 0.001)  # uninformative precision

  for (i in 1:80) # for each of the 80 birds sampled
  {
    mean[i] <- base + a[location[i]] + b[period[i]]  # the mean depends on the location and period
    Y[i] ~ dnorm(mean[i], tau)
  }
}
",file=here("JAGS_mods","box6.4.txt"))
res6.4<-jags(starling2[,c("Y","location","period")],
  model.file=here("JAGS_mods","box6.4.txt"),
  parameters.to.save=c("base","a","b","tau"),n.chains=2,n.iter=110000,n.burnin=10000,n.thin=10)
round(res6.4$BUGSoutput$summary,2)
#DIC comparison
dictab<-data.frame(Model=c("No interaction","Interaction"),
 DIC=round(c(res6.4$BUGSoutput$DIC,res6.5$BUGSoutput$DIC),2),
 pD=round(c(res6.4$BUGSoutput$pD,res6.5$BUGSoutput$pD),2)
)
dictab$deltaDIC<-dictab$DIC-min(dictab$DIC)
dictab
```

## Two way interactions using model matrix and Stan

```{r}
matrixStarling2<-model.matrix(starlinglm2)
starlingList<-list(N=nrow(matrixStarling2),
  Ncoef=ncol(matrixStarling2),
  Y=starling2$Y,
  xMatrix=matrixStarling2
)
stanmatrix1<-stan(file=here("STAN_docs","box5.3matrix.stan"),
  data=starlingList)
print(stanmatrix1,pars=c("b","Sigma"))
```

## Starlings as random effect.

```{r}
write("model   {  # random effects one-way ANOVA
   a ~ dnorm(0, 1.0E-6)  # a is the global mean mass
  for (i in 1:4)  # the deviation from the global mean of the mean mass at each location 
  {
    d[i] ~ dnorm(0.0, tau_a)
  }
  sd_among ~ dunif(0, 100)  #variation among locations
  tau_a <- 1/(sd_among*sd_among)  # convert sd to precision
  sd_within ~ dunif(0, 100)  #  variation between starlings within locations
  tau_w <- 1/(sd_within*sd_within)  # convert sd to precision
  for (i in 1:40) # for each of the 40 birds sampled
  {
    mean[i] <- a + d[location[i]]  # the mean depends on the location
    Y[i] ~ dnorm(mean[i], tau_w)  # data drawn from normal dist'n
  }}
",file=here("JAGS_mods","box6.3.txt"))
res6.3<-jags(starling,model.file=here("JAGS_mods","box6.3.txt"),
  parameters.to.save=c("a","d","sd_among","sd_within"),
  n.chains=2,n.iter=110000,n.burnin=10000,n.thin=10)
round(res6.3$BUGSoutput$summary[c("a","d[1]","d[2]","d[3]","d[4]","sd_among","sd_within"),],2)

dictab<-data.frame(Model=c("random","null"),
  DIC=round(c(res6.3$BUGSoutput$DIC,res6.1null$BUGSoutput$DIC),2),
  pD=round(c(res6.3$BUGSoutput$pD,res6.1null$BUGSoutput$pD),2)) 
dictab$deltaDIC<-dictab$DIC-min(dictab$DIC)
dictab

gg3<-ggs(as.mcmc(res6.3))
gg3<-filter(gg3,Parameter %in% c("d[1]","d[2]","d[3]","d[4]"))
ggs_caterpillar(gg3)

```

## Sleep study random effect

```{r}
library(lme4)
ggplot(sleepstudy,aes(x=Days,y=Reaction))+
  stat_smooth(aes(color=Subject,fill=Subject),method="lm")+
  geom_point(aes(color=Subject))
lmer1 <- lmer(Reaction ~ Days + (Days | Subject),data= sleepstudy)
summary(lmer1)
```

Same in Stan

```{r}
write("
data {
  int<lower=0> N;
  int<lower=0> Nsubject;
  vector[N] y;
  int subject[N];
  vector[N] x;
}
parameters {
  real interceptSigma;
  real slopeSigma;
  real<lower=0> errorSigma;
  real alpha[Nsubject];
  real beta[Nsubject];
}
model {
  errorSigma ~ lognormal(0,10);
  interceptSigma ~ lognormal(0,10);
  slopeSigma ~ lognormal(0,10);
  alpha~normal(0,interceptSigma);
  beta~normal(0,slopeSigma);
  for(i in 1:N) {
   y ~ normal(alpha[subject[i]]+x[i]*beta[subject[i]], errorSigma);
  }
}

  ",file=here("STAN_docs","randomEffects1.stan"))
sleepList<-list(x=sleepstudy$Days,
                y=sleepstudy$Reaction,
                subject=as.numeric(factor(sleepstudy$Subject)),
                N=nrow(sleepstudy),
                Nsubject=length(unique(sleepstudy$Subject)))
sleep1<-stan(file=here("STAN_docs","randomEffects1.stan"),
             data=sleepList)
print(sleep1)
```

```
