---
title: "14. Random effects and INLA"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

To use INLA follow instructions here https://www.r-inla.org/download-install tp to install INLA and here for inlautils https://github.com/timcdlucas/INLAutils
You must install INLAutils with devtools

```{r, message=FALSE}
#devtools::install_github('timcdlucas/INLAutils')
library(INLA)
library(R2jags)
library(tidyverse)
library(ggmcmc)
library(rstan)
library(loo)
library(here)
theme_set(theme_bw())
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

## Sleep study random effect

First, the likelihood version in lme4

```{r}
library(lme4)
sleepstudy<-mutate(sleepstudy,Reaction=Reaction/1000)
ggplot(sleepstudy,aes(x=Days,y=Reaction))+
  stat_smooth(aes(color=Subject,fill=Subject),method="lm",se=FALSE)+
  geom_point(aes(color=Subject))
coef(lm(Reaction~Days,data=sleepstudy))
lmer1 <- lmer(Reaction ~ Days + (1 | Subject),data= sleepstudy)
summary(lmer1)
lmer2 <- lmer(Reaction ~ Days + (Days | Subject),data= sleepstudy)
summary(lmer2)
anova(lmer2,lmer1)  #To get AIC and BIC for model comparison
```

Stan with no random effect (regular regression)

```{r}
write("
data {
  int<lower=0> N;
  vector[N] y;
  vector[N] x;
}
parameters {
  real<lower=0> errorSigma;
  real alpha;
  real beta;
}
model {
  errorSigma ~ exponential(0.01);
  alpha ~ normal(0,100);
  beta~normal(0,100);
  for(i in 1:N) {
   y[i] ~ normal(alpha+x[i]*beta, errorSigma);
  }
}
generated quantities {
  real LL[N];
  for(i in 1:N) {
   LL[i] = normal_lpdf(y[i] | alpha+x[i]*beta, errorSigma);
  }
}

",file="regressionOnly1.stan")
sleepList0<-list(x=sleepstudy$Days,
                y=sleepstudy$Reaction,
                N=nrow(sleepstudy))
sleep0<-stan(file="regressionOnly1.stan",
  data=sleepList0)
print(sleep0,pars=c("alpha","beta","errorSigma"))
ggsSleep0<-ggs(sleep0)
ggs_caterpillar(filter(ggsSleep0,!grepl("LL",Parameter)),sort=FALSE)

```

Same in Stan, with random effect on intercept only

```{r}
write("
data {
  int<lower=0> N;
  int<lower=0> Nsubject;
  vector[N] y;
  int subject[N];
  vector[N] x;
}
parameters {
  real<lower=0> interceptSigma;
  real<lower=0> errorSigma;
  real interceptMu;
  real alpha[Nsubject];
  real beta;
}
model {
  errorSigma ~ exponential(0.01);
  interceptSigma ~ exponential(0.01);
  interceptMu ~ normal(0,100);
  alpha~normal(interceptMu,interceptSigma);
  beta~normal(0,100);
  for(i in 1:N) {
   y[i] ~ normal(alpha[subject[i]]+x[i]*beta, errorSigma);
  }
}
generated quantities {
  real LL[N];
  for(i in 1:N) {
   LL[i] = normal_lpdf(y[i] | alpha[subject[i]]+x[i]*beta, errorSigma);
  }
}

",file="randomEffects1.stan")

sleepList<-list(x=sleepstudy$Days,
                y=sleepstudy$Reaction,
                subject=as.numeric(factor(sleepstudy$Subject)),
                N=nrow(sleepstudy),
                Nsubject=length(unique(sleepstudy$Subject)))

sleep1<-stan(file="randomEffects1.stan",
             data=sleepList,refresh=0)
print(sleep1,pars=c("interceptSigma","interceptMu","errorSigma","beta"))
ggsSleep1<-ggs(sleep1)
ggs_caterpillar(filter(ggsSleep1,!grepl("LL",Parameter)),sort=FALSE)
```

Stan with random effect in both slope and intercept.

```{r}
write("
data {
  int<lower=0> N;
  int<lower=0> Nsubject;
  vector[N] y;
  int subject[N];
  vector[N] x;
}
parameters {
  real<lower=0> interceptSigma;
  real<lower=0> slopeSigma;
  real<lower=0> errorSigma;
  real interceptMu;
  real slopeMu;
  real alpha[Nsubject];
  real beta[Nsubject];
}
model {
  errorSigma ~ exponential(0.01);
  interceptSigma ~ exponential(0.01);
  slopeSigma ~ exponential(0.01);
  interceptMu ~ normal(0,100);
  slopeMu ~ normal(0,100);
  alpha~normal(interceptMu,interceptSigma);
  beta~normal(slopeMu,slopeSigma);
  for(i in 1:N) {
   y[i] ~ normal(alpha[subject[i]]+x[i]*beta[subject[i]], errorSigma);
  }
}
generated quantities {
  real LL[N];
  for(i in 1:N) {
   LL[i] = normal_lpdf(y[i] | alpha[subject[i]]+x[i]*beta[subject[i]], errorSigma);
  }
}

  ",file="randomEffects2.stan")

sleep2<-stan(file="randomEffects2.stan",
             data=sleepList,refresh=0)
print(sleep2,par=c("interceptSigma","errorSigma","slopeSigma",
  "interceptMu","slopeMu"))
ggsSleep2<-ggs(sleep2)
ggs_caterpillar(filter(ggsSleep2,!grepl("LL",Parameter)),sort=FALSE)
```




```{r}
LL0<-extract_log_lik(sleep0,parameter_name = "LL")
LL1<-extract_log_lik(sleep1,parameter_name = "LL")
LL2<-extract_log_lik(sleep2,parameter_name = "LL")
modtab<-data.frame(`Random effects`=0:2,
  waic=c(waic(LL0)$estimate[3,1],
  waic(LL1)$estimate[3,1],
  waic(LL2)$estimate[3,1]))%>%
  mutate(deltaWAIC=waic-min(waic))
round(modtab,2)
```

# INLA

INLA is an efficient alternative to MCMC

```{r}
#Look at  INLA help files
#inla.list.models()
#?inla
#?f
#inla.doc("iid")
```

See here for the sleep model in inla

https://becarioprecario.bitbucket.io/inla-gitbook/ch-multilevel.html

This is a fixed effects regression

```{r,warning=FALSE}
sleepInla0<-inla(Reaction~Days, 
  data=sleepstudy)
summary(sleepInla0)
autoplot(sleepInla0)
```
To pull out model elements

```{r}
sleepInla0$summary.fixed
head(sleepInla0$marginals.fixed$Days)
sleepInla0 $summary.hyperpar
head(sleepInla0$marginals.hyperpar$`Precision for the Gaussian observations`)

sdfunc<-function(x) 1/sqrt(exp(x))
sdMarg<-inla.tmarginal(sdfunc,sleepInla0$internal.marginals.hyperpar[[1]])
head(sdMarg)
ggplot(data.frame(sdMarg),aes(x=x,y=y))+
  geom_line()+
  xlab("Standard deviation")+ylab("Density")

```

To get credible intervals

```{r}
sleepInla0<-inla(Reaction~Days, 
  data=sleepstudy,
  control.predictor = list(compute=TRUE,link=1))
sleep0.fit<-data.frame(sleepInla0$summary.fitted.values) %>%
  bind_cols(sleepstudy)
ggplot(sleep0.fit,aes(x=Days,y=mean))+
    geom_point(aes(y=Reaction))+
    geom_line()+
    geom_ribbon(aes(ymin=X0.025quant,ymax=X0.975quant),alpha=0.2)

```

Residual plots

```{r}
sleep0.fit$residual=sleep0.fit$Reaction-sleep0.fit$mean
ggplot(sleep0.fit,aes(x=mean,y=residual))+
  geom_point()+
  geom_abline(intercept=0,slope=0)

```

Predict at new points

```{r}
new.sleep<-bind_rows(select(sleepstudy,Days,Reaction),
  data.frame(Days=1:10,Reaction=NA))

sleepInla0new<-inla(Reaction~Days, 
  data=new.sleep,
  control.predictor = list(compute=TRUE,link=1))
```

Get WAIC and DIC

```{r}
sleepInla0<-inla(Reaction~Days, 
  data=sleepstudy,
  control.predictor = list(compute=TRUE,link=1),
  control.compute=list(dic=TRUE,waic=TRUE))
summary(sleepInla0)

```

Look at and change priors for fixed effect coefficients

```{r}
#sleep0$all.hyper to look at the priors

#Change precision of fixed effect priors with control.fixed
sleep0.1<-inla(Reaction~Days, 
      control.fixed=list(mean=0,prec=1E-6,mean.intercept=0,
      prec.intercept=1E-6),
  data=sleepstudy)
#sleep0.1$all.hyper
summary(sleep0.1)
```

Check the prior for the variance (which is considered a hyperparameter)
To change prior for variance, must define it in control.family

```{r}
inla.models()$latent$iid$hyper$theta
names(inla.models()$latent$iid$hyper)
sleep0.2<-inla(Reaction~Days,
  control.family=list(hyper=list(prec=list(prior="loggamma",
     param=c(.001,.001)))),
     control.fixed=list(mean=0,prec=1E-6,mean.intercept=0,prec.intercept=1E-6),
      data=sleepstudy)
#sleep0.2$all.hyper                
summary(sleep0.2)
sleep0.2$summary.hyperpar

```

Adding a random intercept

```{r}
sleepInla1<-inla(Reaction~Days+f(Subject,model="iid"), 
  data=sleepstudy,
  control.predictor = list(compute=TRUE,link=1),
  control.compute=list(dic=TRUE,waic=TRUE))
summary(sleepInla1)
autoplot(sleepInla1)
```

Add random intercept and slope

```{r}
sleepstudy$Subject2<-sleepstudy$Subject
sleepInla2<-inla(Reaction ~ Days + f(Subject,model="iid")+
    f(Subject2, Days, model = "iid"),
  data = sleepstudy, 
  control.predictor = list(compute = TRUE),
  control.compute=list(dic=TRUE,waic=TRUE))
summary(sleepInla2)
autoplot(sleepInla2)
```

Model comparison

```{r}
waicInla<-data.frame(random.effect=0:2,
  waic=c(sleepInla0$waic$waic,
  sleepInla1$waic$waic,
  sleepInla2$waic$waic))%>%
  mutate(deltawaic=waic-min(waic))
round(waicInla,2)
```

Fit and residuals

```{r}
sleep2.fit<-data.frame(sleepInla2$summary.fitted.values) %>%
  bind_cols(sleepstudy)%>%
  mutate(residual=Reaction-mean)
ggplot(sleep2.fit,aes(x=Days,y=mean,color=Subject,fill=Subject))+
    geom_point(aes(y=Reaction))+
    geom_line()+
    geom_ribbon(aes(ymin=X0.025quant,ymax=X0.975quant),alpha=0.2)
ggplot(sleep2.fit,aes(x=mean,y=residual))+
    geom_point()+
    geom_abline(intercept=0,slope=0)

```