---
title: "15. Inla and time series"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

```{r, message=FALSE}
#devtools::install_github('timcdlucas/INLAutils')
library(INLAutils)
library(INLA)
library(R2jags)
library(tidyverse)
library(ggmcmc)
library(rstan)
library(loo)
library(lme4)
library(gridExtra)
theme_set(theme_bw())
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

Look at and change priors for fixed effect coefficients

```{r}
sleepInla0<-inla(Reaction~Days, 
  data=sleepstudy)
#sleep0$all.hyper to look at the priors

#Change precision of fixed effect priors with control.fixed
sleep0.1<-inla(Reaction~Days, 
      control.fixed=list(mean=0,prec=1E-6,mean.intercept=0,
      prec.intercept=1E-6),
  data=sleepstudy)
#sleep0.1$all.hyper
summary(sleep0.1)
```

Check the prior for the variance (which is considered a hyperparameter)
To change prior for variance, must define it in control.family

```{r}
inla.models()$latent$iid$hyper$theta
names(inla.models()$latent$iid$hyper)
sleep0.2<-inla(Reaction~Days,
  control.family=list(hyper=list(prec=list(prior="loggamma",
     param=c(.001,.001)))),
     control.fixed=list(mean=0,prec=1E-6,mean.intercept=0,prec.intercept=1E-6),
      data=sleepstudy)
#sleep0.2$all.hyper                
summary(sleep0.2)
sleep0.2$summary.hyperpar

```


## More diagnostics on sleepstudy

```{r}
sleepstudy$Subject2<-sleepstudy$Subject

sleepInla0<-inla(Reaction~Days, 
  data=sleepstudy,
  control.predictor = list(compute = TRUE),
  control.compute=list(dic=TRUE,waic=TRUE,cpo=TRUE))
sleepInla1<-inla(Reaction~Days+f(Subject,model="iid"), 
  data=sleepstudy,
  control.predictor = list(compute = TRUE),
  control.compute=list(dic=TRUE,waic=TRUE,cpo=TRUE))
sleepInla2<-inla(Reaction ~ Days + f(Subject,model="iid")+
    f(Subject2, Days, model = "iid"),
  data = sleepstudy, 
  control.predictor = list(compute = TRUE),
  control.compute=list(dic=TRUE,waic=TRUE,cpo=TRUE))
#check adequacy
summary(sleepInla0$cpo$failure)
summary(sleepInla1$cpo$failure)
summary(sleepInla2$cpo$failure)

#Make table
icInla<-data.frame(random.effect=0:2,
  waic=c(sleepInla0$waic$waic,
  sleepInla1$waic$waic,
  sleepInla2$waic$waic),
  dic=c(sleepInla0$dic$dic,
  sleepInla1$dic$dic,
  sleepInla2$dic$dic),
  looic=-2*c(sum(log(sleepInla0$cpo$cpo)),
  sum(log(sleepInla1$cpo$cpo)),
  sum(log(sleepInla2$cpo$cpo))))%>%
  mutate(deltawaic=waic-min(waic),
    deltadic=dic-min(dic),
    deltalooic=looic-min(looic))
round(icInla,2)

#Make dataframe for plots
sleep0.fit<-sleepstudy %>%
  mutate(fit=sleepInla0$summary.fitted$mean,
    cpo=sleepInla0$cpo$cpo,
    pit=sleepInla0$cpo$pit)
sleep2.fit<-sleepstudy %>%
  mutate(fit=sleepInla2$summary.fitted$mean,
    cpo=sleepInla2$cpo$cpo,
    pit=sleepInla2$cpo$pit)
#CPO
g1<-ggplot(sleep0.fit,aes(x=fit,y=cpo))+
  geom_point()+ggtitle("Simple regression")
g2<-ggplot(sleep2.fit,aes(x=fit,y=cpo))+
  geom_point()+ggtitle("Random intercept and slope")
grid.arrange(g1,g2,ncol=2)
#PIT
g1<-ggplot(sleep0.fit,aes(x=fit,y=pit))+
  geom_point()+
  ggtitle("Simple regression")+
  geom_abline(intercept=0.5,slope=0)
g2<-ggplot(sleep2.fit,aes(x=fit,y=pit))+
  ggtitle("Random intercept and slope")+
  geom_point()+
  geom_abline(intercept=0.5,slope=0)
grid.arrange(g1,g2,ncol=2)
```


## model with logistic regression

```{r}
lr<-read.csv(here("data", "lr.csv"))%>% mutate(across(numrock:nearsouth,~(.x-mean(.x))/sd(.x)))

#INLA version, with default priors for simplicity,adding cpo and pit
lr.inla<-inla(lr~numrock+volcanic+rockcell+nearslope+I(nearslope^2)+nearsouth,
              family="binomial",
              data=lr,
              control.compute=list(cpo=TRUE),
              control.predictor = list(compute=TRUE))
summary(lr.inla)
```

## Put cpo and pit in a dataframe and plot


```{r}
lr.cpo<-cbind(lr,data.frame(lr.inla$cpo))
lr.cpo$Prediction<-lr.inla$summary.fitted.values$mode
g1<-ggplot(lr.cpo,aes(x=Prediction,y=cpo,color=factor(lr)))+geom_point()+
  ggtitle("CPO against Prediction")   
g2<-ggplot(lr.cpo,aes(x=Prediction,y=Prediction-lr,color=factor(lr)))+
  geom_point()+
  ggtitle("Residuals against Prediction")+geom_abline(intercept = 0,slope=0)
grid.arrange(g1,g2)
g1<-ggplot(lr.cpo,aes(x=Prediction,y=pit))+
  geom_point(aes(color=factor(lr)))+
  ggtitle("PIT against prediction") +
  geom_abline(intercept=0.5,slope=0)
g2<-ggplot(lr.cpo,aes(sample=pit))+
  ggtitle("PIT QQ uniform plot") +
  geom_qq(distribution = stats::qunif)+
  geom_qq_line(distribution = stats::qunif)
grid.arrange(g1,g2)
```

## Hierarchical models

## Poisson and Poisson with extra variance

From box 3.4 and 3.5 in McCarthy. First in JAGS

```{r}
write("model
{
  for (i in 1:10)					
  {
    y[i] ~ dpois(lambda)
  }
  lambda ~ dlnorm(0, 1.0E-6)  			
}
",file=here("JAGS_mods", "poisson1.txt"))

data3.4<-data.frame(y=c(6,0,1,2,1,7,1,5,2,0))

poisson1jags<-jags(data=data3.4,
  model.file=here("JAGS_mods", "poisson1.txt"),
  parameters.to.save="lambda")
print(poisson1jags)

write("model
{
  for (i in 1:10)					
  {
    lambda[i] ~ dlnorm(m, tau) 
    y[i] ~ dpois(lambda[i])  	
  }

  m ~ dnorm(0, 1.0E-6)  			
  sd ~ dunif(0, 10)  				
  tau <- 1 / (sd*sd)  				
					
}
",file=here("JAGS_mods", "poisson2.txt"))

poisson2jags<-jags(data=data3.4,
  model.file=here("JAGS_mods", "poisson2.txt"),
  parameters.to.save=c("lambda","m","sd"),
  n.iter=10000)
print(poisson2jags)

```

Now INLA

```{r}
#inla.doc("poisson")
poisson3.4<-inla(formula=y~1,family="poisson",data=data3.4)
round(poisson3.4$summary.fixed,3)
plot(poisson3.4)

#Note that this is on the log scale because of the log link, to transform
poisson3.4$marginals.fixed[[1]]
meanval<-inla.tmarginal(exp,poisson3.4$marginals.fixed[[1]])
plot(meanval)
inla.zmarginal(meanval)
```

## Add quadrat number to the dataframe, for random effect

```{r}
data3.4$quadrat<-1:10
data3.4
poisson3.5<-inla(formula=y~f(quadrat,model="iid"),family="poisson",data=data3.4)
summary(poisson3.5)
plot(poisson3.5)
```

Try again with a more restrictive prior to get more similar to JAGS

```{r}
poisson3.5<-inla(formula=y~f(quadrat,model="iid",
   hyper=list(prec=list(prior="loggamma",param=c(0.01,0.01)
    ,initial=1))),family="poisson",data=data3.4)

round(poisson3.5$summary.fixed,3)
summary(poisson3.5)
plot(poisson3.5)
```

Get quadrat predictions

```{r}
poisson3.5<-inla(formula=y~f(quadrat,model="iid",
   hyper=list(prec=list(prior="loggamma",param=c(0.1,0.1)
    ,initial=1))),family="poisson",data=data3.4,
  control.predictor = list(compute=TRUE))
poisson3.5pred<-data.frame(poisson3.5$summary.fitted.values)
round(poisson3.5pred,2)
```

## Epil example OpenBUGS

Here is the inla version https://sites.google.com/a/r-inla.org/www/examples/volume-1/code-for-epil-example

the Epil dataset loads with the INLA library. 

```{r}
data(Epil)
head(Epil)
summary(Epil)
ggplot(Epil,aes(x=Base,y=y,pch=factor(V4),color=Age))+geom_point()+
  facet_wrap(factor(Trt)~.,ncol=2)
```

Set up in JAGS

```{r}
write("model 
	{
		for(j in 1 : N) {
			for(k in 1 : T) {
				log(mu[j, k]) <- a0 + alpha.Base * (log.Base4[j] - log.Base4.bar)   
	                  + alpha.Trt * (Trt[j] - Trt.bar)  
	                  + alpha.BT  * (BT[j] - BT.bar)  
	                  + alpha.Age * (log.Age[j] - log.Age.bar)  
	                  + alpha.V4  * (V4[k] - V4.bar) 
	                  + b1[j] + b[j, k]
				y[j, k] ~ dpois(mu[j, k])
				b[j, k] ~ dnorm(0.0, tau.b);       # subject*visit random effects
			}
			b1[j]  ~ dnorm(0.0, tau.b1)        # subject random effects
			BT[j] <- Trt[j] * log.Base4[j]    # interaction
			log.Base4[j] <- log(Base[j] / 4) log.Age[j] <- log(Age[j])
		}
		
	# covariate means:
		log.Age.bar <- mean(log.Age[])                
		Trt.bar  <- mean(Trt[])                   
		BT.bar <- mean(BT[])                 
		log.Base4.bar <- mean(log.Base4[])         
		V4.bar <- mean(V4[])                  
	# priors:
	
		a0 ~ dnorm(0.0,1.0E-4) 		           
		alpha.Base ~ dnorm(0.0,1.0E-4)            
		alpha.Trt  ~ dnorm(0.0,1.0E-4);           
		alpha.BT   ~ dnorm(0.0,1.0E-4)            
		alpha.Age  ~ dnorm(0.0,1.0E-4)            
		alpha.V4   ~ dnorm(0.0,1.0E-4)
		tau.b1     ~ dgamma(1.0E-3,1.0E-3); sigma.b1 <- 1.0 / sqrt(tau.b1)
		tau.b      ~ dgamma(1.0E-3,1.0E-3); sigma.b  <- 1.0/  sqrt(tau.b)		     
		        
	# re-calculate intercept on original scale: 
		alpha0 <- a0 - alpha.Base * log.Base4.bar - alpha.Trt * Trt.bar 
		- alpha.BT * BT.bar - alpha.Age * log.Age.bar - alpha.V4 * V4.bar
	}",here("JAGS_mods","epilJags.txt"))


epilWide<-Epil %>% select(-rand,-V4) %>%
  mutate(visit=rep(1:4,max(Ind))) %>% 
  pivot_wider(names_from=visit,values_from=y)
epilList<-list(Trt=epilWide$Trt,
  Base=epilWide$Base,
  Age=epilWide$Age,
  y=epilWide[,5:8],
  V4=c(0,0,0,1),
  N=nrow(epilWide),
  T=4)
epilJags<-jags(data=epilList,
  model.file=here("JAGS_mods","epilJags.txt"),
  parameters.to.save=c("a0","alpha.Base","alpha.Trt","alpha.BT","alpha.Age","alpha.V4","tau.b1","tau.b"))
print(epilJags)
```


Set up in INLA

```{r}
Epil<-Epil %>% mutate(
  CTrt = Trt-mean(Trt),
  ClBase4 = log(Base/4)-mean(log(Base/4)),
  CV4 = V4-mean(V4),
  ClAge = log(Age)-mean(log(Age)),
  CBT = CTrt*ClBase4)
  
formula = y ~ ClBase4 + CTrt + CBT+ ClAge + CV4 +
          f(Ind, model="iid") + f(rand,model="iid")

epil.inla <- inla(formula,family="poisson", data = Epil)
summary(epil.inla)
autoplot(epil.inla)
round(epil.inla$summary.fixed,3)
round(epil.inla$summary.hyperpar,3)
```

Provide improved estimates for the hyperparameters

```{r}
?inla.hyperpar
epil.hyper <- inla.hyperpar(epil.inla)
summary(epil.hyper)
```

Make predictions

```{r}
epil.inla2<-inla(formula,family="poisson", data = Epil, 
                 control.predictor = list(compute=TRUE),
                 control.compute=list(cpo=TRUE,waic=TRUE,dic=TRUE))
pred.epil<-epil.inla2$summary.fitted.values
pred.epil<-bind_cols(pred.epil,Epil)
```


Plot residuals 

```{r}
pred.epil$resid<-pred.epil$y-pred.epil$mode
g1<-ggplot(pred.epil,aes(x=mode,y=resid))+geom_point()+geom_abline(intercept=0,slope=0)
g2<-ggplot(pred.epil,aes(sample=resid))+geom_qq()+geom_qq_line()
grid.arrange(g1,g2,ncol=2)
```


Plot CPO and PIT

```{r}
pred.epil<-bind_cols(pred.epil,data.frame(epil.inla2$cpo))
g1<-ggplot(pred.epil,aes(x=mode,y=cpo))+
  geom_point()+
  ggtitle("CPO against  prediction")
g2<-ggplot(pred.epil,aes(x=mode,y=resid))+
  geom_point()+
  ggtitle("Residuals against Prediction")+
  geom_abline(intercept = 0,slope=0)
g3<-ggplot(pred.epil,aes(x=mode,y=pit))+
  geom_point()+
  ggtitle("PIT against prediction") +
  geom_abline(intercept=0.5,slope=0)
g4<-ggplot(pred.epil,aes(sample=pit))+
  ggtitle("PIT QQ uniform plot") +
  geom_qq(distribution = stats::qunif)+
  geom_qq_line(distribution = stats::qunif)
grid.arrange(g1,g2,g3,g4)
```

## Try a negative binomial model on the same data, removing extra variance term (rand)

```{r}
formula2 <- y ~ ClBase4 + CTrt + CBT+ ClAge + CV4 +
  f(Ind, model="iid") 

#inla.list.models()
#inla.doc("nbinomial")
#inla.doc("nbinomial2")
epil.nb1<-inla(formula2,family="nbinomial", data = Epil, 
                           control.predictor =list(compute=TRUE),
  control.compute=list(cpo=TRUE,waic=TRUE,dic=TRUE))
summary(epil.nb1)
autoplot(epil.nb1)
round(epil.nb1$summary.fixed,2)
round(epil.nb1$summary.hyper,2)
pred.nb<-epil.nb1$summary.fitted.values
pred.nb<-bind_cols(pred.nb,Epil)
pred.nb$resid<-pred.nb$y-pred.nb$mode
g1<-ggplot(pred.nb,aes(x=mean,y=resid))+
  geom_point()+geom_abline(intercept=0,slope=0)
g2<-ggplot(pred.nb,aes(sample=resid))+
  geom_qq()+geom_qq_line()
grid.arrange(g1,g2,ncol=2)

#Plot CPO and PIT
pred.epil.nb<-bind_rows(pred.nb,data.frame(epil.nb1$cpo))
g1<-ggplot(pred.epil,aes(x=mode,y=cpo))+
  geom_point()+
  ggtitle("CPO against  prediction")
g2<-ggplot(pred.epil,aes(x=log(mode),y=resid))+
  geom_point()+
  ggtitle("Residuals against Prediction")+
  geom_abline(intercept = 0,slope=0) 
g3<-ggplot(pred.epil,aes(x=log(mode),y=pit))+geom_point()+
  ggtitle("PIT against prediction")+   
  geom_abline(intercept=0.5,slope=0)
g4<-ggplot(pred.epil,aes(sample=pit))+
  ggtitle("PIT QQ uniform plot") +
  geom_qq(distribution = stats::qunif)+
  geom_qq_line(distribution = stats::qunif)
grid.arrange(g1,g2,g3,g4)
```

## Negative binomial2

```{r}
epil.nb2<-inla(formula2,family="nbinomial2", data = Epil, 
               control.predictor = list(compute=TRUE),
               control.compute=list(waic=TRUE,dic=TRUE,cpo=TRUE))
summary(epil.nb2)
```

# Compare fixed effect parameters

```{r}
x1<-epil.inla$summary.fixed %>%
  rownames_to_column(var = "Coefficient")
x2<-epil.nb1$summary.fixed %>%
  rownames_to_column(var = "Coefficient")
x3<-epil.nb2$summary.fixed %>%
  rownames_to_column(var = "Coefficient")
  
fixedDF<-bind_rows(list(Poisson=x1, Negbin1=x2, Negbin2=x3),.id = "Model")
ggplot(fixedDF,aes(x=Coefficient,y=mean,col=Model))+
  geom_point(position=position_dodge(width=0.2))+
  geom_errorbar(aes(ymin=`0.025quant`,ymax=`0.975quant`),
    width=0.25,lwd=1,position=position_dodge(width=0.2))
```

#Information criteria

```{r}
modtab<-data.frame(Model=c("Poisson","Negbin1","Negbin2"),             
  WAIC=c(epil.inla2$waic$waic,
        epil.nb1$waic$waic,
        epil.nb2$waic$waic),
      DIC=c(epil.inla2$dic$dic,
        epil.nb1$dic$dic,
        epil.nb2$dic$dic))
modtab$deltaDIC=modtab$DIC-min(modtab$DIC)
modtab$deltaWAIC=modtab$WAIC-min(modtab$WAIC)

modtab
```

## Zero inflated poisson

```{r}
epil.zip0<-inla(formula,family="zeroinflatedpoisson0", data = Epil, 
               control.compute = list(dic = TRUE, waic = TRUE,cpo=TRUE))

summary(epil.zip0)

epil.zip1<-inla(formula,family="zeroinflatedpoisson1", data = Epil, 
               control.compute = list(dic = TRUE, waic = TRUE,cpo=TRUE))
summary(epil.zip1)
autoplot(epil.zip1)
```

Model comparison

```{r}
modtab<-data.frame(Model=c("Poisson","ZIP0","ZIP1"),
                   WAIC=c(epil.inla2$waic$waic,epil.zip0$waic$waic,
                   epil.zip1$waic$waic),
                   DIC=c(epil.inla2$dic$dic,epil.zip0$dic$dic,epil.zip1$dic$dic))
modtab$deltaDIC=modtab$DIC-min(modtab$DIC)
modtab$deltaWAIC=modtab$WAIC-min(modtab$WAIC)
modtab

```

