---
title: "23. Spatial models"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---


```{r,message=FALSE}
library(tidyverse)
library(R2jags)
library(ggmcmc)
library(kableExtra)
library(INLA)
library(INLAutils)
library(SpatialEpi)
library(inlabru)
library(geoR) #for matern function
library(here)
theme_set(theme_bw())
```

## Spatial models. Conditional auto-regression

The Scotland lip cancer example CAR

```{r}
data(Scotland)
summary(Scotland)
#load the map
g <- system.file("demodata/scotland.graph", package="INLA")
g
scotland.matrix<-inla.graph2matrix(g)
head(scotland.matrix)

#For the CAR model
formula <- Counts~I(X/10)+f(Region,model="besag", graph = g,
                   hyper=list(prec=list(param=c(0.5,0.0005)) ))

mod.scotland <- inla(formula,family="poisson",E=E,data=Scotland,
                    control.compute=list(waic=TRUE,dic=TRUE))

summary(mod.scotland)
autoplot(mod.scotland)
```

For a simple random effects model

```{r}
head(Scotland)
formulaR <- Counts~I(X/10)+f(Region,model="iid",
                           hyper=list(prec=list(param=c(0.5,0.0005)) ))

mod.scotlandR <- inla(formulaR,family="poisson",E=E,data=Scotland,
                     control.compute=list(waic=TRUE,dic=TRUE))
summary(mod.scotlandR)
```

Incorporating both a random effect and a CAR model, Called BYM model or convolution prior

```{r}
inla.doc("bym")

formulaBYM<- Counts~I(X/10)+f(Region,model="bym",graph=g,
                              hyper=list(prec.unstruct=list(param=c(0.5,0.0005)),
                                         prec.spatial=list(param=c(0.5,0.0005))))

mod.scotlandBYM <- inla(formulaBYM,family="poisson",E=E,data=Scotland,
                    control.compute=list(waic=TRUE,dic=TRUE))
summary(mod.scotlandBYM)
```

Model comparison

```{r}
modtab<-data.frame(model=c("CAR","Random","BYM"),
                   DIC=c(mod.scotland$dic$dic,mod.scotlandR$dic$dic,mod.scotlandBYM$dic$dic),
                   WAIC=c(mod.scotland$waic$waic,mod.scotlandR$waic$waic,mod.scotlandBYM$waic$waic))
modtab$deltaDIC<-modtab$DIC-min(modtab$DIC)
modtab$deltaWAIC<-modtab$WAIC-min(modtab$WAIC)
modtab
```

Relative risk is just the fitted values, since INLA automatically includes the link, 
but not the offset, when using E rather than offset. 

```{r}
RR<-mod.scotland$summary.fitted.values[,"mean"]
Scotland$RR<-RR
```

Plot observed vs. expected counts and make maps

```{r}
ggplot(Scotland,aes(x=Counts,y=E*RR))+
  geom_point()+
  geom_abline(intercept=0,slope=1)

## Make maps. 
data(scotland)
scotland.map<-scotland$spatial.polygon
plot(scotland.map)

#Must have same row names in the polygon file and the data frame to be able to map a variable
row.names(Scotland)
names(scotland.map)
row.names(Scotland)<-names(scotland.map)
counties<-SpatialPolygonsDataFrame(scotland.map,Scotland)
#Map of model estimated relative risk
redblue<-colorRampPalette(c("blue","white","red"))
spplot(counties,zcol="RR",main="Relative Risk",col.regions=redblue(100))
#Map of original counts
greenblue<-colorRampPalette(c("blue","white","green"))
spplot(counties,zcol="Counts",main="Raw counts",col.regions=greenblue(100))

```

CPO and PIT

```{r}
mod.scotland <- inla(formula,family="poisson",E=E,data=Scotland,
                    control.compute=list(waic=TRUE,dic=TRUE,cpo=TRUE))

Scotland$cpo<-mod.scotland$cpo$cpo
Scotland$pit<-mod.scotland$cpo$pit
head(Scotland)
ggplot(Scotland,aes(x=cpo))+geom_histogram()
counties<-SpatialPolygonsDataFrame(scotland.map,Scotland)
spplot(counties,zcol="cpo",main="CPO",col.regions=greenblue(100))

ggplot(Scotland,aes(x=pit))+geom_histogram()+ggtitle("PIT")
spplot(counties,zcol="pit",main="PIT",col.regions=greenblue(100))
```


## Geostatistical models

Plot Matern function

```{r}
df<-expand.grid(Distance=seq(0.1,5,.1),nu=seq(0.5,2,0.5),kappa=seq(.5,1,0.5))
df$Correlation<-matern(df$Distance,df$nu,df$kappa)
ggplot(df,aes(x=Distance,y=Correlation,color=factor(nu)))+
  geom_line(lwd=1)+theme_bw()+facet_wrap(kappa~.,ncol=1)
```

This code is simplified from Blangiardo and Cameletti's book, chapter 6. See BlangiardoChapter6.r for the complete code

SPDE toy data example
```{r}
data(SPDEtoy)
dim(SPDEtoy)
head(SPDEtoy)
ggplot(SPDEtoy,aes(x=s1,y=s2,size=y))+geom_point(alpha=0.5)
```

Make meshes

```{r}
?inla.mesh.2d
coords <- as.matrix(SPDEtoy[,1:2])
mesh0 <- inla.mesh.2d(loc=coords, max.edge=0.1)
ggplot(SPDEtoy,aes(x=s1,y=s2))+
  gg(mesh0)+
  geom_point() 

mesh1 <- inla.mesh.2d(loc=coords, max.edge=c(0.1, 0.1))
ggplot(SPDEtoy,aes(x=s1,y=s2))+
  gg(mesh1)+
  geom_point() 

mesh2 <- inla.mesh.2d(loc=coords, max.edge=c(0.1, 0.2))
ggplot(SPDEtoy,aes(x=s1,y=s2))+
  gg(mesh2)+
  geom_point() 

mesh3 <- inla.mesh.2d(loc=coords, max.edge=c(0.1, 0.2), offset=c(0.4,0.1))
ggplot(SPDEtoy,aes(x=s1,y=s2))+
  gg(mesh3)+  
  geom_point() 

mesh4 <- inla.mesh.2d(loc=coords, max.edge=c(0.1, 0.2), offset=c(0.1,0.4))
ggplot(SPDEtoy,aes(x=s1,y=s2))+
  gg(mesh4)+  
  geom_point() 

#With a spatial domain
domain <- matrix(cbind(c(0,1,1,0.7,0), c(0,0,0.7,1,1)),ncol=2)
mesh6 <- inla.mesh.2d(loc.domain=domain, max.edge=c(0.04, 0.2), cutoff=0.05, offset = c(0.1, 0.4))
ggplot(SPDEtoy,aes(x=s1,y=s2))+
  gg(mesh6)+  
  geom_point() 
```


Make projection matrix, so we can map the output to the mesh

```{r}
A.est1 <- inla.spde.make.A(mesh=mesh1, loc=coords)
dim(A.est1)
head(A.est1)
table(as.vector(A.est1))
A.est6 <- inla.spde.make.A(mesh=mesh6, loc=coords)
dim(A.est6)
```

Set up correlation structure for spde

```{r}
inla.spde.models()
?inla.spde2.matern
spde <- inla.spde2.matern(mesh=mesh6)
summary(spde)
spde$n.theta
```

Set up formula and run INLA
Data should be a list, because it requires inputs of different lengths (data points vs. mesh vertices)

```{r}
dataList <- list(y=SPDEtoy$y, intercept=rep(1,spde$n.spde),spatial.field=1:spde$n.spde)
summary(dataList)
```

The spatial distribution of Y is estimated at the vertices, not the data points.

Suppress the default intercept with -1 and add an intercept we will set up in the data input

```{r}
formula <- y ~ -1 + intercept + f(spatial.field, model=spde)

#Add A to the control.predictor list to generated fitted values at the data points
output6 <- inla(formula,
                data=dataList,
                control.predictor=list(A=A.est6,compute=TRUE))

round(output6$summary.fixed,3)
round(output6$summary.hyperpar[1,],3)
summary(output6)

output6.field <- inla.spde2.result(inla=output6, name="spatial.field", 
  spde=spde, do.transf=TRUE)
summary(output6.field)

# Summary of parameters
inla.zmarginal(output6.field$marginals.kappa[[1]])
inla.zmarginal(output6.field$marginals.variance.nominal[[1]])
inla.zmarginal(output6.field$marginals.range.nominal[[1]])

#Credible interval from marginal
inla.hpdmarginal(0.95, output6.field$marginals.kappa[[1]])
inla.hpdmarginal(0.95, output6.field$marginals.variance.nominal[[1]])
inla.hpdmarginal(0.95, output6.field$marginals.range.nominal[[1]])
```

Use inla.stack to set up design matrix and other data inputs

```{r}
?inla.stack
?inla.spde.make.index #Extracts necessary indices for spatial field, and other random effects
s.index <- inla.spde.make.index(name="spatial.field", n.spde=spde$n.spde)
summary(s.index)

#data= the y data as a list
#A = the projection matrix as a list
#effects = list of variables to include in this model at the knots
stack.est <- inla.stack(data=list(y=SPDEtoy$y),
                        A=list(A.est6),
                        effects=list(c(s.index, list(intercept=1))), 
                        tag="est") #Estimation
summary(stack.est)
output6.stack <- inla(formula,
                      data=inla.stack.data(stack.est, spde=spde),
                      family="gaussian",
                      control.predictor=list(A=inla.stack.A(stack.est), 
                                             compute=TRUE))             

summary(output6.stack)
summary(output6)
```

Results with stack are identical

Generate spatial predictors on a grid, rather than on the mesh

```{r}
grid.x <- 50
grid.y <- 50
pred.grid <- expand.grid(x = seq(0, 1, length.out = grid.x), y = seq(0, 1, length.out = grid.y))
dim(pred.grid)

A.pred6 <- inla.spde.make.A(mesh=mesh6, loc=as.matrix(pred.grid))
dim(A.pred6)

stack.pred.response <- inla.stack(data=list(y=NA), 
                                  A=list(A.pred6),
                                  effects=list(c(s.index, list(intercept=1))),
                                  tag="pred.response") 

join.stack <- inla.stack(stack.est, stack.pred.response)

join.output <- inla(formula,
                    data=inla.stack.data(join.stack),
                    control.predictor=list(A=inla.stack.A(join.stack), compute=TRUE))
```

Getting predictions to plot

```{r}
pred.grid$index <- inla.stack.index(join.stack, tag="pred.response")$data
pred.grid$mean <- join.output$summary.fitted.values[pred.grid$index,"mean"]
pred.grid$sd <- join.output$summary.fitted.values[pred.grid$index,"sd"]

ggplot(pred.grid,aes(x=x,y=y,fill=mean))+
  geom_tile() 
ggplot(pred.grid,aes(x=x,y=y,fill=sd))+
  geom_tile() 
```

Gambia malaria example from B and C, chapter 6

```{r}
data(gambia)
head(gambia)

coords <-  as.matrix(gambia[,1:2])/1000 #in km
summary(gambia)
dim(gambia)

# Create an index at the village level
ind <- paste("x",coords[,1], "y", coords[,2], sep="")
# Detect non duplicated villages (Have the same x and y coordinates)
which.nodupl <- which(!duplicated(ind))
village.index <- c(NA, length=nrow(gambia))
village.index[1 : (which.nodupl[length(which.nodupl)]-1)] <-
  rep(1:64,times=as.numeric(diff(which.nodupl)))
village.index[which.nodupl[length(which.nodupl)] : nrow(gambia)] <- 65
gambia$village.index <- village.index

summary(gambia)
```

Make non-convex hull around villages

```{r}
bnd <- inla.nonconvex.hull(coords,convex=-0.1)
gambia.mesh <- inla.mesh.2d(boundary = bnd,offset=c(30, 60), max.edge=c(20,40))
ggplot(gambia,aes(x=x/1000,y=y/1000))+
  gg(gambia.mesh)+  
  geom_point() 
```

Set up spde and A matrix

```{r}
gambia.spde <- inla.spde2.matern(mesh=gambia.mesh)
summary(gambia.spde)
gambia.spde$n.spde
gambia.spde$n.theta
A.est <- inla.spde.make.A(mesh=gambia.mesh, loc=coords)
dim(A.est)
s.index <- inla.spde.make.index(name="spatial.field",n.spde=gambia.spde$n.spde)
summary(s.index$spatial.field.group)
gambia.stack.est <- inla.stack(data=list(y=gambia$pos),
                               A=list(A.est, 1, 1, 1, 1, 1, 1),
                               effects=
                                 list(c(s.index, list(Intercept=1)),
                                      list(age=gambia$age/365),
                                      list(treated=gambia$treated),
                                      list(netuse=gambia$netuse),
                                      list(green=gambia$green),
                                      list(phc=gambia$phc),
                                      list(village.index=gambia$village.index)),
                               tag="est")

formula <- y ~ -1 + Intercept + treated + netuse + age + green + phc + 
  f(spatial.field, model=gambia.spde) + f(village.index, model="iid")

gambia.output <- inla(formula,
                      data=inla.stack.data(gambia.stack.est, spde=gambia.spde),
                      family="binomial",Ntrials=1,
                      control.predictor=list(A=inla.stack.A(gambia.stack.est), compute=TRUE),
                      control.compute=list(dic=TRUE,waic=TRUE))

fixed.out <- round(gambia.output$summary.fixed[,1:5],3)
fixed.out
autoplot(gambia.output)
summary(gambia.output)
gambia.output$summary.hyperpar
```

Precision for village.index

```{r}
sigma2u_marg <- inla.tmarginal(function(x) 1/x, gambia.output$marginals.hyperpar[[3]])
sigma2u_m1 <- inla.emarginal(function(x) x, sigma2u_marg)
sigma2u_m2 <- inla.emarginal(function(x) x^2, sigma2u_marg)
sigma2u_stdev <- sqrt(sigma2u_m2 - sigma2u_m1^2)
sigma2u_quantiles <- inla.qmarginal(c(0.025, 0.5, 0.975), sigma2u_marg)


mod.field <- inla.spde2.result(gambia.output, name="spatial.field", gambia.spde)
summary(mod.field)

inla.emarginal(function(x) exp(x)/(1+exp(x)),gambia.output$marginals.fixed[["netuse"]])
inla.emarginal(inla.link.invlogit, gambia.output$marginals.fixed[["netuse"]])
```

Model WITHOUT the spatial effect

```{r}
gambia.stack.est.noGF <- inla.stack(data=list(y=gambia$pos),
                                    A=list(1, 1, 1, 1, 1, 1, 1),
                                    effects=
                                      list(list(Intercept=rep(1,nrow(gambia))),
                                           list(age=gambia$age/365),
                                           list(treated=gambia$treated),
                                           list(netuse=gambia$netuse),
                                           list(green=gambia$green),
                                           list(phc=gambia$phc),
                                           list(village.index=gambia$village.index)),
                                    tag="est")

formulanoGF <- y ~ -1 + Intercept + treated + netuse + age + green + phc +
  f(village.index,model="iid")

gambia.output.noGF <- inla(formulanoGF,
                           data=inla.stack.data(gambia.stack.est.noGF, spde=gambia.spde),
                           family="binomial",Ntrials=1,
                           control.predictor=list(A=inla.stack.A(gambia.stack.est.noGF), compute=TRUE),
                           control.compute=list(dic=TRUE,waic=TRUE))


modtab<-data.frame(Model=c("Spatial","No spatial"),
                   DIC=c(gambia.output$dic$dic,gambia.output.noGF$dic$dic),
                   WAIC=c(gambia.output$waic$waic,gambia.output.noGF$waic$waic))
modtab$deltaWAIC=modtab$WAIC-min(modtab$WAIC)
modtab$deltaDIC=modtab$DIC-min(modtab$DIC)
modtab
```

