---
title: "24. Ecological models"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

## sdmTMB for spatial and spatiotemporal models

```{r,message=FALSE}
library(sdmTMB)
library(tidyverse)
library(R2jags)
library(ggmcmc)
library(gridExtra)
library(here)
theme_set(theme_bw())
ilogit<-function(x) 1/(1+exp(-x))
```

This is the Pacific cod example from here
https://pbs-assess.github.io/sdmTMB/articles/basic-intro.html

Make mesh 
```{r}
mesh <- make_mesh(pcod, c("X", "Y"), cutoff = 10)
plot(mesh)
```

Run model with matern correlation and check diagnostics

```{r}
cod1 <- sdmTMB(
  data = pcod,
  formula = present ~ depth_scaled + depth_scaled2,
  mesh = mesh,
  family = binomial(link = "logit"),
  spatial = "on"
)
sanity(cod1)
tidy(cod1, conf.int = TRUE)
tidy(cod1, "ran_pars", conf.int = TRUE)
pcod$residual<-residuals(cod1)
ggplot(pcod, aes(X, Y, col = residual)) +
  scale_colour_gradient2() +
  geom_point() 
?residuals.sdmTMB
```

Make predictions on a new grid and plot

```{r}
predictions <- predict(cod1, newdata = qcs_grid)
plot_map <- function(dat, column) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    coord_fixed()
}
g1<-plot_map(predictions, ilogit(est)) +
  scale_fill_viridis_c(na.value = "yellow", 
                       limits = c(0, quantile(ilogit(predictions$est), 0.995))) +
  theme(legend.position="bottom")+
   ggtitle("Prediction (fixed effects + all random effects)")
g2<-plot_map(predictions, ilogit(est_non_rf)) +
  scale_fill_viridis_c()+theme(legend.position="bottom") +
  ggtitle("Prediction (fixed effects only)")
g3<- plot_map(predictions, omega_s) +
  scale_fill_gradient2() +
  theme(legend.position="bottom")+
  ggtitle("Spatial random effects only")
grid.arrange(g1,g2,g3,ncol=3)
```

Add temporal effect, where year is random effect

```{r}
cod2 <- sdmTMB(
  data = pcod,
  formula = present ~ depth_scaled + depth_scaled2,
  mesh = mesh,
  family = binomial(link = "logit"),
  spatial = "on",
  time = "year",
  spatiotemporal = "IID"
)
sanity(cod2)
tidy(cod2, conf.int = TRUE)
tidy(cod2, "ran_pars", conf.int = TRUE)
qcs_grid2<-NULL
for(i in 1:length(unique(pcod$year))) qcs_grid2<-bind_rows(qcs_grid2,qcs_grid)
qcs_grid2$year<-rep(unique(pcod$year),each=nrow(qcs_grid))
predictions <- predict(cod2, newdata = qcs_grid2)
plot_map(predictions, ilogit(est)) +
  scale_fill_viridis_c(
    # trim extreme high values to make spatial variation more visible
    na.value = "yellow", limits = c(0, quantile(ilogit(predictions$est), 0.995))) +
  facet_wrap(~year) +
  ggtitle("Prediction (fixed effects + all random effects)")

```
## Multivariate models
## Multinomial with no predictors koala. 

```{r}

write("model{
  Y[1:9] ~ dmulti(p[1:9], N)     # N is equal to sum(Y[])
  # number of koalas in each tooth wear class drawn from a multinomial distribution
  p[1:9] ~ ddirch(alpha[])
  # uninformative prior for proportions (p[]) if all values of alpha are equal to one
}
",file=here("JAGS_mods","koalaMulti.txt"))
koala<-list(N=397, Y=c(55, 132, 88, 48, 31, 26, 14, 3, 0), alpha=c(1, 1, 1, 1, 1, 1, 1, 1, 1))
init1<-list(list(p=rep(1/9,9)),list(p=rep(1/9,9)))
koalaRes1<-jags(koala,init1,parameters.to.save = c("p"),
          model.file=here("JAGS_mods","koalaMulti.txt"),n.chains = 2,n.iter = 10000,n.burnin = 1000)
round(koalaRes1$BUGSoutput$summary,2)
```

Same in Poisson multinomial format

```{r}
write("model{
  alpha[1]<-0
  for(i in 2:9) {alpha[i]~dnorm(0,1.0E-6)}
  lambda~dunif(-1000,1000) #sample size parameter
  for(i in 1:9) {
    log(mu[i])<-lambda+alpha[i]
    Y[i]~dpois(mu[i])
    p[i]<-mu[i]/sum(mu[])
  }
}
",file=here("JAGS_mods","koalaPoissonMulti.txt"))
init1<-list(list(lambda=1,alpha=c(NA,3,4,3,1,1,.1,.1,-1)),
            list(lambda=2,alpha=c(NA,3,4,3,1,1,.1,.1,-3)))
koala2<-list(Y=c(55, 132, 88, 48, 31, 26, 14, 3, 0))
koalaRes2<-jags(koala2,init1,parameters.to.save = c("p","alpha","lambda","mu"),
               model.file=here("JAGS_mods","koalaPoissonMulti.txt"),n.chains = 2,n.iter = 10000,n.burnin = 1000)
round(koalaRes2$BUGSoutput$summary,3)
```

## Simulated fin data and poisson multinomial regression for Field et al. example

Simulated data with 2 random effects, like Hong Kong fin data  

```{r}
p=c(0.35,0.25,0.2,0.1,0.1)  # True average proportion of each species
n=rpois(20,1000)      # Sample size
x1=rep(1:5,4)     # First predictor 
x2=rep(1:4,each=5)   # 2nd predicotr
lp=log(p/(1-p)) #Logit scale average proportion

# Random effects are generated for each level of each factor, for each species
x1.eff=matrix(rnorm(25,0,.01),5,5)
x2.eff=matrix(rnorm(25,0,.01),5,5)
pmat=matrix(0,20,5)  
lpmat=matrix(0,20,5)  #Logit of proportion of each species in each sample
for(i in 1:20) 
  for(j in 1:5) {
    lpmat[i,j]=rnorm(1,lp[j]+x1.eff[x1[i],j]+x2.eff[x2[i],j],.05)  #additive predictors
  }
pmat=exp(lpmat)/(1+exp(lpmat)) # Convert logit to probabilities and adjust to sum to 1
sum(pmat)
pmat=round(pmat,3)
sum(pmat)
for(i in 1:20) pmat[i,5]=1-sum(pmat[i,1:4])
summary(as.vector(pmat))   # Make sure all probabilites are positive. 
Y=pmat  # Use multinomial to generate counts
for(i in 1:20)
  Y[i,]=rmultinom(1,n[i],pmat[i,])
Y

## Make input data for jags models
dat.ran=list(Y=Y,N=dim(Y)[1],K=5,x1=x1,x2=x2,n1=max(x1),n2=max(x2))
write("# Nuisance parameter applies to first species only. 
model {
 #Priors
    beta[1]<-0  #Species effect
    for(k in 2:K) {  beta[k]~dnorm(0,1.0E-6)  }
    tau1~dgamma(0.001,0.001)  #Random effects
    tau2~dgamma(0.001,0.001)
    for(i in 1:n1) { gamma[i,1]<-0 }
    for(i in 1:n2) { delta[i,1]<-0 }
    for(k in 2:K) { #Loop over species k
     for(i in 1:n1) { gamma[i,k]~dnorm(0,tau1) }
     for(i in 1:n2) { delta[i,k]~dnorm(0,tau2) }
    }
  for(i in 1:N) {   lambda[i]~dunif(-1000,1000) }
# Poisson multinomial likelihood
  for(i in 1:N)  {   #Loop over samples
  for (k in 1 : K) {     # loop over species
   Y[i, k] ~ dpois(mu[i,k])
   log(mu[i, k]) <- beta[k]+gamma[x1[i],k]+delta[x2[i],k]+lambda[i]
   p[i,k]<-mu[i,k]/sum(mu[i,])
 }}  
 for(k in 2:K) {
     gamma.new[k]~dnorm(0,tau1)
     delta.new[k]~dnorm(0,tau2)
}
 gamma.new[1]<-0
 delta.new[1]<-0
 for(k in 1:K) {
     new.mu[k]<-exp(beta[k]+gamma.new[k]+delta.new[k])
     Ppred[k]<-new.mu[k]/sum(new.mu[])
     fixed.mu[k]<-exp(beta[k])
     Pfix[k]<-fixed.mu[k]/sum(fixed.mu[])		
 }  
}

",file=here("JAGS_mods","finsPoissonNoInterceptContrast.txt"))
init1=list(beta=c(NA,0.1,0.1,0.1,0.1),tau1=1,tau2=1)
init2=list(beta=c(NA,0,0,0,0),tau1=2,tau2=2)

# Jags model using Poisson-multinomial
res.ran1<-jags(dat.ran,list(init1,init2),parameters.to.save=c("p","beta","gamma","Pfix","Ppred","tau1","tau2","new.mu","fixed.mu","lambda"),
             model.file=here("JAGS_mods","finsPoissonNoInterceptContrast.txt"),n.iter=10000,n.thin=1,n.burnin=1000,
             n.chains=2)

## Make graph of predicted proportions
res<-res.ran1$BUGSoutput
p.row<-paste("p[",c(rep(1:20,5)),",",rep(1:5,each=20),"]",sep="")
p.res<-data.frame(res$summary[p.row,])
p.res$p.true<-as.vector(pmat)
p.fix<-data.frame(res$summary[paste0("Pfix[",1:5,"]"),])
p.fix$p.true<-p
ggplot(p.res,aes(x=p.true,y=mean))+geom_point()+geom_abline()+ 
  geom_errorbar(aes(ymin=X2.5.,ymax=X97.5.),width=0.01)+
  geom_errorbar(data=p.fix,aes(x=p.true,y=mean,ymin=X2.5.,ymax=X97.5.),col="red",lwd=2,width=0.01)+
  xlab("True proportion")  +ylab("Estimated proportion")
```


## Correlation from Box 5.11 in McCarthy

```{r}
write("model
{
  mean[1] ~ dnorm(0.0, 1.0E-6)
  mean[2] ~ dnorm(0.0, 1.0E-6)

  sd[1] ~ dunif(0, 100)
  sd[2] ~ dunif(0, 100)
  correl ~ dunif(-1, 1)

  covar[1,1] <- sd[1]*sd[1]
  covar[2,2] <- sd[2]*sd[2]
  covar[1,2] <- correl*sd[1]*sd[2]
  covar[2,1] <- correl*sd[1]*sd[2]

  omega[1 : 2 , 1 : 2]  <- inverse(covar[ , ])

  for (i in 1:N)
  {
    Y[i, 1:2] ~ dmnorm(mean[], omega[ , ])
  }
}
",file=here("JAGS_mods","box5.11.txt"))


dat1<-list(N=8, Y=matrix(c(2.15, 39, 2.27, 38, 4.31, 61, 2.58, 79, 3.23, 35, 1.83, 39, 1.54, 45, 2.00, 28),8,2,byrow=TRUE))
dat1$Y

inits<-list(list(mean = c(2, 2), sd = c(.1, .1), correl = 0.2),list(mean = c(1, 1), sd = c(1, 1), correl = 0))

rescor<-jags(dat1,inits,parameters.to.save = c("mean","sd","correl","covar"),
  model.file=here("JAGS_mods","box5.11.txt"), n.chains = 2,n.iter = 100000,n.thin=10,n.burnin = 10000)
round(rescor$BUGSoutput$summary,2)
```

## Jaws example from Lunn. Multivariate normal. 

```{r}
write("  model {
    for (i in 1:20) {Y[i, 1:4] ~ dmnorm(mu[], Sigma.inv[,])}
    for (j in 1:4)  {mu[j]    <- alpha + beta*x[j]}
    alpha                      ~ dnorm(0, 0.0001)
    beta                       ~ dnorm(0, 0.0001)
    Sigma.inv[1:4, 1:4]        ~ dwish(R[,], 4)
    Sigma[1:4, 1:4]           <- inverse(Sigma.inv[,])
  }
",file=here("JAGS_mods","model6.4.1.txt"))

#Data converted from original bugs format
dat=list(Y = structure(
           .Data = c(47.8, 48.8, 49.0, 49.7,
                     46.4, 47.3, 47.7, 48.4,
                     46.3, 46.8, 47.8, 48.5,
                     45.1, 45.3, 46.1, 47.2,
                     47.6, 48.5, 48.9, 49.3,
                     52.5, 53.2, 53.3, 53.7,
                     51.2, 53.0, 54.3, 54.5,
                     49.8, 50.0, 50.3, 52.7,
                     48.1, 50.8, 52.3, 54.4,
                     45.0, 47.0, 47.3, 48.3,
                     51.2, 51.4, 51.6, 51.9,
                     48.5, 49.2, 53.0, 55.5,
                     52.1, 52.8, 53.7, 55.0,
                     48.2, 48.9, 49.3, 49.8,
                     49.6, 50.4, 51.2, 51.8,
                     50.7, 51.7, 52.7, 53.3,
                     47.2, 47.7, 48.4, 49.5,
                     53.3, 54.6, 55.1, 55.3,
                     46.2, 47.5, 48.1, 48.4,
                     46.3, 47.6, 51.3, 51.8),
           .Dim =  c(4, 20)),
       x = c(8.0, 8.5, 9.0, 9.5),
       R = structure(
           .Data = c(4, 0, 0, 0,
                     0, 4, 0, 0,
                     0, 0, 4, 0,
                     0, 0, 0, 4),
           .Dim =  c(4, 4)))
dat$Y=t(dat$Y)  #Note to convert from bugs to R, reverse the dimensions then transpose
names(dat)
dat$R
dat$Y
dat$x
inits=list(list(alpha=20,beta=0),list(alpha=30,beta=1))

res1<-jags(dat,inits,parameters.to.save = c("alpha","beta","mu","Sigma"),
  model.file = here("JAGS_mods","model6.4.1.txt"),n.chains = 2,n.iter = 10000,n.burnin = 1000)

round(res1$BUGSoutput$summary,2)
sigmat<-matrix(res1$BUGSoutput$summary[1:16,"mean"],4,4)
round(sigmat,2)

#Old R plotting because of the matrix format
matplot(dat$x,t(dat$Y),type="p",pch=1,col=1,xlab="Age",ylab="Jaw length",bty="n")
matlines(dat$x,res1$BUGSoutput$summary[paste0("mu[",1:4,"]"),c("2.5%","50%","97.5%")],lty=c(2,1,2),col=1)
for(i in 1:20) lines(dat$x,dat$Y[i,],col=i)
```

## Ramirez et al. turtle meta-analysis

See model10E.r

## Run example data in siber to estimate ellipses.  
See examples in https://cran.r-project.org/web/packages/SIBER/vignettes/Introduction-to-SIBER.html

```{r}
library(SIBER)

data("demo.siber.data")
siber.example <- createSiberObject(demo.siber.data)
summary(siber.example$original.data)
community.hulls.args <- list(col = 1, lty = 1, lwd = 1)
group.ellipses.args  <- list(n = 100, p.interval = 0.95, lty = 1, lwd = 2)
group.hull.args      <- list(lty = 2, col = "grey20")

par(mfrow=c(1,1))
plotSiberObject(siber.example,
                ax.pad = 2, 
                hulls = F, community.hulls.args, 
                ellipses = T, group.ellipses.args,
                group.hulls = T, group.hull.args,
                bty = "L",
                iso.order = c(1,2),
                xlab = expression({delta}^13*C~'\u2030'),
                ylab = expression({delta}^15*N~'\u2030')
)


group.ML <- groupMetricsML(siber.example)
print(group.ML)
```

Run Bayesian model

```{r}
parms <- list()
parms$n.iter <- 2 * 10^4   # number of iterations to run the model for
parms$n.burnin <- 1 * 10^3 # discard the first set of values
parms$n.thin <- 10     # thin the posterior by this many
parms$n.chains <- 2        # run this many chains

# define the priors
priors <- list()
priors$R <- 1 * diag(2)
priors$k <- 2
priors$tau.mu <- 1.0E-3

?siberMVN
ellipses.posterior <- siberMVN(siber.example, parms, priors)
summary(ellipses.posterior[[1]])
SEA.B <- siberEllipses(ellipses.posterior)

siberDensityPlot(SEA.B, xticklabels = colnames(group.ML), 
                 xlab = c("Community | Group"),
                 ylab = expression("Standard Ellipse Area " ('\u2030' ^2) ),
                 bty = "L",
                 las = 1,
                 main = "SIBER ellipses on each group"
)
```

## Normal mixture model eyes from Lunn 11.6.1

```{r}
write("model {
  for (i in 1:N) {
    y[i]     ~ dnorm(mu[i], tau)
    mu[i]   <- lambda[T[i]]
    T[i]     ~ dcat(p[])
  }
  p[1:2]     ~ ddirch(alpha[])
  alpha[1]  <- 1
  alpha[2]  <- 1
  theta      ~ dunif(0, 1000)
  lambda[2] <- lambda[1] + theta
  lambda[1]  ~ dunif(-1000, 1000)
  sigma      ~ dunif(0, 100)
  tau       <- 1 / pow(sigma, 2)
 
 # generate a prediction from this model
  T.pred     ~ dcat(p[])	
  y.pred     ~ dnorm(lambda[T.pred], tau)
}",here("JAGS_mods","model11.6.1.txt"))

dat4=list(y = c(529.0, 530.0, 532.0, 533.1, 533.4, 533.6, 533.7, 534.1, 534.8, 535.3,
                535.4, 535.9, 536.1, 536.3, 536.4, 536.6, 537.0, 537.4, 537.5, 538.3,
                538.5, 538.6, 539.4, 539.6, 540.4, 540.8, 542.0, 542.8, 543.0, 543.5,
                543.8, 543.9, 545.3, 546.2, 548.8, 548.7, 548.9, 549.0, 549.4, 549.9,
                550.6, 551.2, 551.4, 551.5, 551.6, 552.8, 552.9,553.2), N = 48,
          T = c(1, NA, NA, NA, NA, NA, NA, NA, NA, NA,
                NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,			
                NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
                NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,
                NA, NA, NA, NA, NA, NA, NA, 2))

inits4=list(list(lambda = c(535, NA), theta = 5, sigma = 10),
            list(lambda = c(435, NA), theta = 3, sigma = 5))


res4=jags(dat4,inits4,parameters.to.save=c("lambda","theta","T.pred","y.pred"),
          model.file=here("JAGS_mods","model11.6.1.txt"),n.iter=100000,n.thin=1,n.burnin=10000,
          n.chains=2)
round(res4$BUGSoutput$summary,2)
plot(density(res4$BUGSoutput$sims.matrix[,"y.pred"]),main="Mixed distribution",lwd=2,col="red",xlab="Eye wavelength")
```

## SIMMR intro https://cran.r-project.org/web/packages/simmr/vignettes/simmr.html

```{r}
library(simmr)
mix <- matrix(c(
  -10.13, -10.72, -11.39, -11.18, -10.81, -10.7, -10.54,
  -10.48, -9.93, -9.37, 11.59, 11.01, 10.59, 10.97, 11.52, 11.89,
  11.73, 10.89, 11.05, 12.3
), ncol = 2, nrow = 10)
colnames(mix) <- c("d13C", "d15N")
s_names <- c("Zostera", "Grass", "U.lactuca", "Enteromorpha")
s_means <- matrix(c(-14, -15.1, -11.03, -14.44, 3.06, 7.05, 13.72, 5.96), ncol = 2, nrow = 4)
s_sds <- matrix(c(0.48, 0.38, 0.48, 0.43, 0.46, 0.39, 0.42, 0.48), ncol = 2, nrow = 4)
c_means <- matrix(c(2.63, 1.59, 3.41, 3.04, 3.28, 2.34, 2.14, 2.36), ncol = 2, nrow = 4)
c_sds <- matrix(c(0.41, 0.44, 0.34, 0.46, 0.46, 0.48, 0.46, 0.66), ncol = 2, nrow = 4)
conc <- matrix(c(0.02, 0.1, 0.12, 0.04, 0.02, 0.1, 0.09, 0.05), ncol = 2, nrow = 4)

simmr_in <- simmr_load(
  mixtures = mix,
  source_names = s_names,
  source_means = s_means,
  source_sds = s_sds,
  correction_means = c_means,
  correction_sds = c_sds,
  concentration_means = conc
)

plot(simmr_in)
?simmr_mcmc
simmr_out <- simmr_mcmc(simmr_in)
summary(simmr_out, type = 'diagnostics')
post_pred <- posterior_predictive(simmr_out)
plot(simmr_out, type = 'density')

compare_sources(simmr_out, 
                source_names = c('Zostera','U.lactuca'))

```
