---
title: "Lecture 6"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---
# R code for JAGS examples and plots from lecture 6

```{r,message=FALSE}
library(R2jags)
library(tidyverse)
library(ggmcmc)
library(gridExtra)
library(here)
theme_set(theme_bw()) #Set black and white theme for ggplot
```

# Powerful owl binomial example  

## Uniform prior on m

```{r}
write("model {
m ~ dunif(0.02, 0.2)
m.prior ~ dunif(0.02, 0.2)
x ~ dbin(m, n)
x.prior.sim ~ dbin(m.prior,n)
x.sim ~ dbin(m,n)
}
",file= here("Jags_mods","Owl.survival.txt"))

owldat<-list(x=1,n=36)
init1<-list(m=0.1)
init2<-list(m=0.2)

owl1<-jags(data=owldat,inits=list(init1,init2),
  parameters.to.save=c("m","m.prior","x.prior.sim","x.sim"),
  model.file=here("Jags_mods","Owl.survival.txt"),
              n.iter=100000,n.burnin=10000,n.thin=1,n.chains=2)
round(owl1$BUGSoutput$summary,3)
ggs_density(filter(ggs(as.mcmc(owl1)),Parameter=="m"))
```

## Uniform prior on average age

```{r}
write("model
{
ave.age ~ dunif(5, 50)
M.inst<- 1/ave.age
m <- 1-exp(-M.inst)
ave.age.prior~ dunif(5, 50)
m.prior <- 1-exp(-1/ave.age.prior)
x ~ dbin(m, n)
x.prior.sim ~ dbin(m.prior,n)
x.sim ~ dbin(m,n)
}
",file=here("Jags_mods","Owl.M.txt"))
init1<-list(ave.age=15)
init2<-list(ave.age=20)

owl2<-jags(owldat,list(init1,init2),
  parameters.to.save=c("m","ave.age","ave.age.prior","m.prior","x.prior.sim","x.sim"),
  model.file=here("Jags_mods","Owl.M.txt"),
              n.iter=100000,n.burnin=10000,n.thin=1,n.chains=2)
round(owl2$BUGSoutput$summary,3)
ggs_density(filter(ggs(as.mcmc(owl2)),Parameter!="deviance"))
```

## Informative prior from beta distribution

```{r}
afunc<-function(mu,sig) mu*(mu*(1-mu)/sig^2-1)
bfunc<-function(mu,sig) (1-mu)*(mu*(1-mu)/sig^2-1)
a<-afunc(0.11,0.05)
b<-bfunc(0.11,0.05)
a;b
ggplot(data.frame(x=c(0,1)),aes(x=x))+
    stat_function(fun=dbeta,args=list(shape1=a,shape2=b),linewidth=2)+
    ylab("Propability")+
    ggtitle("Informative Prior")
```

Jags model with informative prior

```{r}
write("
model 
{
m ~ dbeta(4.198, 33.96)
m.prior ~ dbeta(4.198, 33.96)
x ~ dbin(m, n)
x.prior.sim ~ dbin(m.prior,n)
x.sim ~ dbin(m,n)
}
",file=here("Jags_mods","Owl.Info.txt"))

init1<-list(m=0.1)
init2<-list(m=0.2)

owl3<-jags(owldat,list(init1,init2),
  parameters.to.save=c("m","m.prior","x.prior.sim","x.sim"),
  model.file=here("Jags_mods","Owl.Info.txt"),
              n.iter=100000,n.burnin=10000,n.thin=1,n.chains=2)
round(owl3$BUGSoutput$summary,3)
owl3gg<-ggs(as.mcmc(owl3))
ggs_density(filter(owl3gg,Parameter %in% c("m","m.prior")))
```

## To plot owls

```{r}
gg1<-ggs(as.mcmc(owl1))
gg2<-ggs(as.mcmc(owl2))
gg3<-ggs(as.mcmc(owl3))
g1<-ggs_density(filter(gg1,Parameter=="m"))+ggtitle("Prior uniform on M") +xlim(c(0,0.25))
g2<-ggs_density(filter(gg2,Parameter=="m"))+ggtitle("Prior uniform on mean age") +xlim(c(0,0.25))
g3<-ggs_density(filter(gg3,Parameter=="m"))+ggtitle("Informative prior")+xlim(c(0,0.25))
grid.arrange(g1,g2,g3)
```

Plot just the confidence intervals

```{r}
df1<-data.frame(rbind(owl1$BUGSoutput$summary["m",],owl2$BUGSoutput$summary["m",],owl3$BUGSoutput$summary["m",]))
df1$Prior<-c("Uniform on M","Uniform on Age","Informative")
ggplot(df1,aes(x=Prior,y=mean,ymin=X2.5.,ymax=X97.5.))+geom_point()+geom_errorbar(width=0.1)+ylab("Posterior mean and 95% CI")
```

# PMPD with owls. Change data to NA, and set DIC to false
```{r}
owlNodat<-list(x=NA,n=36)
init1<-list(m=0.1)
init2<-list(m=0.2)

owl1pmpd<-jags(owlNodat,list(init1,init2),parameters.to.save=c("m","m.prior"),
  model.file=here("Jags_mods","Owl.survival.txt"),
  n.iter=100000,n.burnin=10000,n.thin=1,
  n.chains=2,
  DIC=FALSE)

init1<-list(ave.age=15)
init2<-list(ave.age=20)

owl2pmpd<-jags(owlNodat,list(init1,init2),parameters.to.save=c("m","m.prior"),model.file=here("Jags_mods","Owl.M.txt"),
              n.iter=100000,n.burnin=10000,n.thin=1,n.chains=2,DIC=FALSE)

init1<-list(m=0.1)
init2<-list(m=0.2)

owl3pmpd<-jags(owlNodat,list(init1,init2),parameters.to.save=c("m","m.prior"),
  model.file=here("Jags_mods","Owl.Info.txt"),
              n.iter=100000,n.burnin=10000,n.thin=1,n.chains=2,DIC=FALSE)

#ggmcmc doesn't work with no data, and only one parameter, so make plots with ggplot

pmpd.df1<-bind_rows(list(`Uniform on M` =data.frame(owl1pmpd$BUGSoutput$sims.matrix),
  `Uniform on ave age`=data.frame(owl2pmpd$BUGSoutput$sims.matrix),
  `Informative`=data.frame(owl2pmpd$BUGSoutput$sims.matrix)),.id="Prior")
ggplot(pmpd.df1) +
  stat_density(aes(x=m,fill="Posterior"),alpha=0.5)+
  stat_density(aes(x=m.prior,fill="Prior"),alpha=0.5)+
  facet_wrap(Prior~.,ncol=1)+
  ggtitle("Prior and posterior with no data")

```

## Posterior and prior predictive distributions

```{r}
sim.df1<-bind_rows(list(`Uniform on M` =data.frame(owl1$BUGSoutput$sims.matrix),
  `Uniform on ave age`=data.frame(owl2$BUGSoutput$sims.matrix),
  `Informative`=data.frame(owl2$BUGSoutput$sims.matrix)),.id="Prior")
ggplot(sim.df1) +
  geom_bar(aes(x=x.sim,fill="Posterior"),alpha=0.5)+
  geom_bar(aes(x=x.prior.sim,fill="Prior"),alpha=0.5)+
  facet_wrap(Prior~.,ncol=1)+
  geom_vline(xintercept=1,color="red")+
  ggtitle("Prior and posterior predictive distributions")+
  xlab("Number of mortalities")


```


# Box 3.4 Poisson mean

```{r}
write("model{
  for (i in 1:10)					# for each of the ten quadrats
  {
     y[i] ~ dpois(m)			# number of trees drawn from Poisson with mean m
  }

  m ~ dlnorm(0.0, 1.0E-6)  # uninformative prior for mean trees per quadrat
}",here("Jags_mods","model3.4.txt"))

Data<-list(y=c(6,0,1,2,1,7,1,5,2,0))

InitialValues<-list(list(m=5),list(m=4))

model3.4.res<-jags(data=Data,
  inits = InitialValues,
  parameters.to.save = c("m"),
  n.chains = 2,
  n.burnin = 1000,
  n.iter = 20000,
  n.thin = 1,
  model.file=here("Jags_mods","model3.4.txt"))

model3.4.res$BUGSoutput
ggs3.4<-filter(ggs(as.mcmc(model3.4.res)),Parameter=="m")
ggs_density(ggs3.4)

#Fit compared to data
ggplot(data.frame(Trees=Data$y),aes(x=Trees))+geom_histogram(bins=8,fill="lightblue",color="darkblue")+
  geom_line(data=data.frame(Trees=0:7,Count=length(Data$y)*dpois(0:7,2.5)),aes(x=Trees,y=Count),lwd=2)
```
  



# Negative binomial count data example 

```{r}
write("model{
  for (i in 1:N)					{
     y[i] ~ dnegbin(p,r)			
  }
  p~dunif(0,1)
  r~dgamma(0.01,0.001)
  m<-r*(1-p)/p
  v<-r*(1-p)/(p*p)
  dispersion<-1/p
  step.prop<-step(dispersion-1)
}
",file=here("Jags_mods","NegBinDefault.txt"))

dat1<-list(y=c(6,0,1,2,1,7,1,5,2,0),N=10)
init1<-list(p=0.5,r=2)
init2<-list(p=0.1,r=3)

negbinres1<-jags(dat1,list(init1,init2),parameters.to.save=c("r","p","m","v",
               "dispersion","step.prop"),model.file=here("Jags_mods","NegBinDefault.txt"),
              n.iter=100000,n.burnin=10000,n.thin=1,n.chains=2)
round(negbinres1$BUGSoutput$summary,2)
negbindf<-ggs(as.mcmc(negbinres1))
ggs_density(filter(negbindf,Parameter=="m"))
```

The mean alternate parameterization of the negative binomial

```{r}
?dnbinom
write("model{
  for (i in 1:N)	  {
     y[i] ~ dnegbin(p,r)			
  }
  r ~dlnorm(0.0, 1.0E-6)
  m~dlnorm(0.0, 1.0E-6)  
  p<-r/(r+m)
  v<-m+m*m/r
  dispersion<-1/p
  step.prop<-step(dispersion-1)
}
",file=here("Jags_mods","NegBinMean.txt"))

dat1<-list(y=c(6,0,1,2,1,7,1,5,2,0),N=10)
init1<-list(m=2,r=2)
init2<-list(m=4,r=3)

negbinres2<-jags(dat1,list(init1,init2),parameters.to.save=c("r","p","m","v",
               "dispersion","step.prop"),model.file=here("Jags_mods","NegBinMean.txt"),
              n.iter=100000,n.burnin=10000,n.thin=1,n.chains=2)
round(negbinres2$BUGSoutput$summary,2)
negbindfe<-ggs(as.mcmc(negbinres1))
ggs_density(filter(negbindfe,Parameter=="m"))
ggs_pairs(filter(negbindf,Parameter %in% c("r","p")),lower=list(continuous="density"))
```

# Koala multinomial example
```{r}
write("
model
{
  Y[1:9] ~ dmulti(p[1:9], N)     
  p[1:9] ~ ddirich(alpha[])  
}
",file=here("Jags_mods","koala.txt"))

dat1<-list(N=397, Y=c(55, 132, 88, 48, 31, 26, 14, 3, 0), 
  alpha=c(1, 1, 1, 1, 1, 1, 1, 1, 1))
init1<-list(p=rep(1/9,9))

koala1<-jags(dat1,list(init1,init1),parameters.to.save=c("p"),
  model.file=here("Jags_mods","koala.txt"),
              n.iter=100000,n.burnin=10000,n.thin=1,n.chains=2)
round(koala1$BUGSoutput$summary,3)
koala1$BUGSoutput$summary["p[1]",]
?ggs
koalaggs<-ggs(as.mcmc(koala1),family="p")
ggs_caterpillar(koalaggs)
ggs_grb(koalaggs)
```

