---
title: "Homework 10"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

```{r,message=FALSE}
library(tidyverse)
library(R2jags)
library(ggmcmc)
library(kableExtra)
theme_set(theme_bw())
```

## 1. Mark-recapture

```{r}
data<-read.csv("dat10.1.csv")
summary(data)
dat1<-list(N=27,Years=4,First=data$First,Y=data[,1:4])
init1<-list(surv=0.9,resight=0.5)
init2<-list(surv=0.8,resight=0.4)

write("model
{
  for (i in 1:N)  # for each bird
  {
    alive[i, First[i]] <- 1  # 1 means it is alive the first time it was seen
    for (j in (First[i]+1):Years)  # for each year after the first
    {
      palive[i,j] <- surv * alive[i, j-1] 
      alive[i,j] ~ dbern(palive[i,j])
      psight[i,j] <- resight * alive[i, j]        
      Y[i, j] ~ dbern(psight[i,j])
    }
}
# Uninformative priors for survival and resighting rates
  surv ~ dunif(0, 1)  # uninformative
  resight ~ dunif(0,1)  # resighting rate - uninformative
}
",file="HW10.1a.txt")

res1a<-jags(dat1,list(init1,init2),parameters.to.save=c("surv","resight"),model.file="HW10.1a.txt",
          n.chains=2,n.iter=10000,n.burnin=1000,n.thin=1)
round(res1a$BUGSoutput$summary,2)
ggs1a<-ggs(as.mcmc(res1a))
ggs_density(ggs1a)
```

Rhat is near 1, and the effective sample size is large, so convergence is not a problem. 

# 1b

```{r}
write("model
{
  for (i in 1:N)  # for each bird
  {
    alive[i, First[i]] <- 1  # 1 means it is alive the first time it was seen
    for (j in (First[i]+1):Years)  # for each year after the first
    {
      palive[i,j] <- surv[j-1] * alive[i, j-1] # Survival is from previous year
      alive[i,j] ~ dbern(palive[i,j])
      psight[i,j] <- resight * alive[i, j]        
      Y[i, j] ~ dbern(psight[i,j])
    }
  }
  for(i in 1:3) { surv[i] ~ dunif(0, 1) } # uninformative
  surv[4]<-surv[3]
  resight~dunif(0,1)
}
",file="HW10.1b.txt")
alive<-as.matrix(data[,1:4])
alive[]<-1
for(i in 1:27) alive[i,1:data$First[i]]<-NA
head(alive)
init1<-list(surv=c(0.9,.9,.9,NA),resight=0.5,alive=alive)
init2<-list(surv=c(0.8,.8,.8,NA),resight=0.4,alive=alive)
res1b<-jags(dat1,list(init1,init2),parameters.to.save=c("surv","resight"),model.file="HW10.1b.txt",
           n.chains=2,n.iter=10000,n.burnin=1000,n.thin=1)
round(res1b$BUGSoutput$summary,2)
ggs1b<-ggs(as.mcmc(res1b))
ggs_density(ggs1b)
```

Convergence looks fine. The survival probabilities are not very different from part a, implying they are fairly consistent between years. 

## 1c
```{r}
write("model
{
  for (i in 1:N)  # for each bird
  {
    alive[i, First[i]] <- 1  # 1 means it is alive the first time it was seen
    for (j in (First[i]+1):Years)  # for each year after the first
    {
      palive[i,j] <- surv[j-1] * alive[i, j-1] 
      alive[i,j] ~ dbern(palive[i,j])
      psight[i,j] <- resight[i] * alive[i, j]        
      Y[i, j] ~ dbern(psight[i,j])
    }
}
# Uninformative priors for survival and resighting rates
  for(i in 1:3) { surv[i] ~ dunif(0, 1) } # uninformative
  surv[4]<-surv[3]
  mu~dnorm(0,0.00001)
  tau~dgamma(0.01,0.01)
  for(i in 1:N) { 
   lresight[i]~dnorm(mu,tau)
   logit(resight[i])<-lresight[i]
#   resight[i]<-exp(lresight[i])/(1+exp(lresight[i]))
  }
}",file="HW10.1c.txt")
init1<-list(surv=c(0.9,.9,.9,NA),tau=1,alive=alive)
init2<-list(surv=c(0.8,.8,.8,NA),tau=0.4,alive=alive)
res1c<-jags(dat1,list(init1,init2),parameters.to.save=c("surv","resight","mu","tau"),model.file="HW10.1c.txt",
           n.chains=2,n.iter=10000,n.burnin=1000,n.thin=1)

round(res1c$BUGSoutput$summary,2)
ggs10.1<-ggs(as.mcmc(res1c))
ggs_caterpillar(ggs10.1,family="resight")
```

There isn't much variation between individuals in resighting probability, so this component of the model may not be needed. Note, I didn't specify whether survival should vary between years, so its fine if you had only one survival parameter. 

## d

```{r}
data.frame(Model=c("A","B","C"),
  DIC=c(res1a$BUGSoutput$DIC,
   res1b$BUGSoutput$DIC,
   res1c$BUGSoutput$DIC)) %>%
  mutate(delta=DIC-min(DIC))%>% kable(digits=2)
```

DIC prefers the model in part b, where survival varies by year, but detection probability does not vary by individual. 

## 2. Missing data

```{r}
swoPres<-read.csv("swoPres.csv")
summary(swoPres)
```

## a. Complete cases

```{r}

write("model{
  a~dnorm(0,1.0E-16)
  b1~dnorm(0,1.0E-16)
  b2~dnorm(0,1.0E-16)
  for(i in 1:N) {
    logit(p[i])<-a+b1*Year[i]+b2*hbf[i]
    Presence[i]~dbern(p[i])
  }
  for(i in 1:26) {
    logit(yearpred[i])<-a+b1*i+b2*mean(hbf[])
  }
}",file="logisticRegression.txt")

swoComplete<-na.omit(swoPres)
summary(swoComplete)
presdat<-list(Year=swoComplete$Year,Presence=swoComplete$Presence,
  hbf=swoComplete$hbf,
  N=nrow(swoComplete))
summary(presdat)
res10.2a<-jags(presdat,
  model.file="logisticRegression.txt",
  parameters.to.save=c("a","b1","b2","yearpred"),
  n.chain=2,n.iter=21000,n.thin=1,n.burnin=1000
  )
range(res10.2a$BUGSoutput$summary[,"n.eff"])
range(res10.2a$BUGSoutput$summary[,"Rhat"])
```

# i. Summary statistics for intercept and slopes

```{r}
round(res10.2a$BUGSoutput$summary[c("a","b1","b2"),],3)
```

# ii. Year effect 

```{r}
ggs2a<-ggs(as.mcmc(res10.2a))
ggs_caterpillar(ggs2a,family="yearpred",sort=FALSE) +coord_flip()
```

## b. with missing y

No need to change the code, just use the data that includes the NA values

```{r}
presdat2<-list(Year=swoPres$Year,Presence=swoPres$Presence,hbf=swoPres$hbf,N=nrow(swoPres))
summary(presdat2)
str(presdat2)
res10.2b<-jags(presdat2,
  model.file="logisticRegression.txt",
  parameters.to.save=c("a","b1","b2","yearpred"),
  n.chain=2,n.iter=21000,n.thin=1,n.burnin=1000
  )
range(res10.2b$BUGSoutput$summary[,"n.eff"])
range(res10.2b$BUGSoutput$summary[,"Rhat"])
```

# i. Summary statistics for intercept and slope

```{r}
round(res10.2b$BUGSoutput$summary[c("a","b1","b2"),],3)
```

# ii. Year effect 

```{r}
ggs2b<-ggs(as.mcmc(res10.2b))
ggs_caterpillar(ggs2b,family="yearpred",sort=FALSE)+coord_flip()
```

## c. 

The results are very similar, so in this case, the missing data filling doesn't seem to make much of a difference. This is a little surprising, since I specifically put in the NAs when when the swordfish were more abundant.  

## 3 Missing  X data

```{r}
# library(ISwR)
# temp<-cystfibr %>%
#   select(pemax, weight,masspercent=bmp,expirevolume=fev1,residvolume=rv) 
# #%>%
# #  mutate(across(weight:residvolume,~(.x-mean(.x))/sd(.x)))
# temp$weight[c(2,3,4,16,18,20,15)]<-NA
# write.csv(temp,file="pemax.csv",row.names = FALSE)
pemax<-read.csv("pemax.csv")
summary(pemax)
```

## a

```{r}
write("model{
  for(i in 1:5) {
    b[i]~dnorm(0,1.0E-6)
  }
  prec~dgamma(0.01,0.001)
  for(i in 1:N) {
   mu[i]<-b[1]+b[2]*weight[i]+b[3]*masspercent[i]+b[4]*expirevolume[i]+b[5]*residvolume[i]
   pemax[i]~dnorm(mu[i],prec)
  }
}
",file="regression.txt")

pemax.complete<-na.omit(pemax)
pemax.complete<-c(N=nrow(pemax.complete),as.list(pemax.complete))

mod10.3a<-jags(pemax.complete,
               parameters.to.save = c("b","prec","mu"),
               model.file = "regression.txt",
               n.iter = 10000,n.burnin = 1000)
round(mod10.3a$BUGSoutput$summary[c("b[1]","b[2]","b[3]","b[4]","b[5]","prec"),],3) 
mod103a<-ggs(as.mcmc(mod10.3a))
ggs_caterpillar(filter(mod103a,Parameter %in% c("b[2]","b[3]","b[4]","b[5]")),sort = FALSE)
```

## b

```{r}
write("model{
  for(i in 1:5) {
    b[i]~dnorm(0,1.0E-6)
  }
  meanwt~dnorm(0,0.001)
  prec~dgamma(0.01,0.001)
  precwt~dgamma(0.01,0.001)
  for(i in 1:N) {
   weight[i]~dnorm(meanwt,precwt)
   mu[i]<-b[1]+b[2]*weight[i]+b[3]*masspercent[i]+b[4]*expirevolume[i]+b[5]*residvolume[i]
   pemax[i]~dnorm(mu[i],prec)
  }
}
",file="regression2.txt")

pemax.list<-c(N=nrow(pemax),as.list(pemax))

mod10.3b<-jags(pemax.list,
               parameters.to.save = c("b","prec","mu","residvolume","meanwt","precwt"),
               model.file = "regression2.txt",
               n.iter = 10000,n.burnin = 1000)

round(mod10.3b$BUGSoutput$summary[c("b[1]","b[2]","b[3]","b[4]","b[5]","prec"),],3) 

round(mod10.3a$BUGSoutput$summary[c("b[1]","b[2]","b[3]","b[4]","b[5]","prec"),],3) 

round(mod10.3b$BUGSoutput$summary[c("b[1]","b[2]","b[3]","b[4]","b[5]","prec","precwt","meanwt"),],3) 
mod103b<-ggs(as.mcmc(mod10.3b))
ggs_caterpillar(filter(mod103b,Parameter %in% c("b[2]","b[3]","b[4]","b[5]")),sort = FALSE)


allcoef<-bind_rows(list(complete=filter(mod103b,Parameter %in% c("b[2]","b[3]","b[4]","b[5]")),                        impute=filter(mod103a,Parameter %in% c("b[1]","b[2]","b[3]","b[4]","b[5]"))),.id="method") %>%
  group_by(Parameter,method)%>%
  summarize(ymin=quantile(value,0.025),
            ymax=quantile(value,0.975),
            ymean=mean(value))

ggplot(allcoef,aes(x=Parameter,y=ymean,ymin=ymin,ymax=ymax,color=method))+
  geom_point(position=position_dodge(width=0.1))+
  geom_errorbar(width=0.1,position=position_dodge(width=0.1))
```
The effects of each regression variable were simiolar but not identical. The credible intervals of the parameters are also not very different. 