---
title: "Homework 1"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(tidyverse)
library(R2jags)
library(ggmcmc)
```


## 1. DeGeorge Syndrome

```{r}
DeGeorge<-data.frame(Disease=c("yes","no"), Prob.disease=c(1/4000,3999/4000),
  Conditional.prob=c(0.99,0.0011))
DeGeorge$Prob.pos.and.disease=DeGeorge$Prob.disease*DeGeorge$Conditional.prob
DeGeorge$Prob.disease.given.positive<-DeGeorge$Prob.pos.and.disease/sum(DeGeorge$Prob.pos.and.disease)
DeGeorge
```

Prior P(disease)=0.00025

P(positive result|disease)=0.99

P(positive result|no disease)=0.0011

By Bayes rule,the probability of having the disease given a positive result is the fraction of the positives that are true rather than false positives
$$P(disease|positive) = \frac{P(positive|disease)P(disease)}{P(positive|disease)P(disease)+P(positive|no disease)P(no disease)}=$$
$$P(disease|positive) = \frac{0.99*0.00025}{0.99*0.00025+0.0111*0.99975)}=$$ `r round(DeGeorge$Prob.disease.given.positive[1],3)`


## 2. Binomial likelihood

```{r}
females<-c(8, 10, 8, 8, 7, 8, 9, 10, 6, 7, 7, 4, 8, 6, 8, 
7, 8, 10, 6, 7)
turtle<-data.frame(females=females,index=1:20)
```

## a. Find the MLE p with a grid search

Since the sample size is 20, we need to loop over all the potential values of the proportion female and calculate the sum of the likelihood across the 20 samples. 

```{r}
sexRatio<-data.frame(p=seq(0.1,0.99,by=0.01))
for(i in 1:nrow(sexRatio)) {
 sexRatio$LL[i]<-sum(dbinom(females,10,sexRatio$p[i],log=TRUE))
}
sexRatio[sexRatio$LL==max(sexRatio$LL),]
```
The MLE of p is 0.76.

## b  Plot the likelihood profile

```{r}
ggplot(sexRatio,aes(x=p,y=LL))+geom_line()+
  ggtitle("Log likelihood")
```
## c. Minimize the negative log likelihood

```{r}
getNegLogLike<-function(p,data) {
  x<-data$females
  -sum(dbinom(x,10,p,log=TRUE))
}

sexRatioModel<-optim(par=0.5,
                      fn=getNegLogLike,
                     method="Brent",
                     lower=0,
                     upper=1,
                     data=turtle
                     )
sexRatioModel$par

```
The MLE value is the same. 

## 3. Normal likelihood

```{r}
lengthDat<-c(26.4, 16.1, 26.8, 26.5, 29, 22.4, 21.4, 25.2, 26.1, 16.3, 24.2, 
20.9, 11.5, 22.2, 22.1, 19.1, 24.4, 24.1, 21.5, 20.2)
data<-data.frame(length=lengthDat)
```

## a. Grid search

```{r}
meanSdVal<-expand.grid(mean=seq(1,30,length=100),sd=seq(0.1,10,length=100))
for(i in 1:nrow(meanSdVal)) {
  meanSdVal$LL[i]<-sum(dnorm(data$length,meanSdVal$mean[i],meanSdVal$sd[i],log=TRUE))
}
meanSdVal[meanSdVal$LL==max(meanSdVal$LL),]
```
The MLE of the mean is 22.4 and sd is 4.1

## b Objective function
```{r}
getNegLogLikeNormal<-function(pars,data) {
  x=data$length
  mu=pars[1]
  sd=pars[2]
  LL<-dnorm(x,mu,sd,log=TRUE)
  -sum(LL)
}
meanSdModel<-optim(par=c(20,2),
                  fn=getNegLogLikeNormal,
                  data=data)
meanSdModel$par
```
The values are identical to what we got with the grid search. 

## c. Hessian

```{r}

meanSdModel<-optim(par=c(20,2),
                  fn=getNegLogLikeNormal,
                  data=data,
                  hessian=TRUE)
solve(meanSdModel$hessian)

```
The inverse of the hessian is the variance covariance function. It shows a very low correlation (0.0013) between the mean and the sd, which is expected.

## d Pointwise plot

```{r}
data$LL<-dnorm(data$length,meanSdModel$par[1],meanSdModel$par[2],log=TRUE)
ggplot(data,aes(x=length,y=LL))+geom_point()
```

Data points that are closer to the mean have higher pointwise likelihood, as
we expect. 