---
title: "HW2"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,message=FALSE}
library(tidyverse)
library(ggmcmc)
library(R2jags)
```


# 1. Calculating means

## a Frequentist mean and its standard deviation

```{r}
weightVals<-c(46, 51, 35, 46, 26, 37, 49, 48, 30, 46, 21, 26, 37, 36, 49, 42, 50, 33, 52, 37)
summary(weightVals)
meanval<-mean(weightVals)
sdval<-sd(weightVals)
seval<-sd(weightVals)/sqrt(length(weightVals))
```

The mean is `r round(meanval,2)` and its standard error is `r round(seval,2)`

## b  Analytic credible interval 

```{r}
priormean<-35
priorsd<-2
posterior.mean<-(priormean/priorsd^2+meanval*20/sdval^2)/(1/priorsd^2+20/sdval^2)
posterior.sd<-1/sqrt(20/sdval^2+1/priorsd^2)
CI<-c(posterior.mean-1.96*posterior.sd,posterior.mean+1.96*posterior.sd)
```

Posterior mean = `r round(posterior.mean, 2)`

Posterior sd = `r round(posterior.sd, 2)`

Credible interval = `r round(CI[1],2)` to `r round(CI[2],2)`

## c Grid search

```{r}
meanGrid<-data.frame(mean=seq(from=21,to=52,by=0.01))
for(i in 1:nrow(meanGrid)) {
  meanGrid$LL[i]<-sum(dnorm(weightVals,meanGrid$mean[i],sdval,log=TRUE))
  meanGrid$logPrior[i]<-dnorm(meanGrid$m[i],priormean,priorsd,log=TRUE)
  meanGrid$logPost[i]<-meanGrid$LL[i]+meanGrid$logPrior[i]
}  
meanGrid$mean[meanGrid$logPost==max(meanGrid$logPost)]
```
The maximum posterior probability estimate is `r round(meanGrid$mean[meanGrid$logPost==max(meanGrid$logPost)],1)`

## d. JAGS MCMC

```{r}
write("model{
  m ~ dnorm(35,0.25)	#  vague prior
  prec <- 1 / ( stdev*stdev)	# precision of the data = 1/variance
  for (i in 1:20)		# for each of the ten trees
  {
    Y[i] ~ dnorm(m, prec) # diameter drawn from normal (likelihood)
  }
}", file="HW1.1.1.txt")

HW1.1.1<-jags(data=list(Y=weightVals,stdev=sd(weightVals)),
  parameters.to.save = "m",
  model.file = "HW1.1.1.txt",
  n.iter=100000,
  n.burnin=10000
)
round(HW1.1.1$BUGSoutput$summary,3)
```

MCMC solution is mean=`r round(HW1.1.1$BUGSoutput$summary["m","mean"],2)`,and sd of the mean =`r round(HW1.1.1$BUGSoutput$summary["m","sd"],2)`.

## e. Compare methods

The values are similar for the analytic, grid search and except for the uncertainty caused by MCMC sampling, as expected. 

## f. With different prior

```{r}
write("model{
  m ~ dnorm(35,1.0E-6)	#  vague prior
  prec <- 1 / ( stdev*stdev)	# precision of the data = 1/variance
  for (i in 1:20)		# for each of the ten trees
  {
    Y[i] ~ dnorm(m, prec) # diameter drawn from normal (likelihood)
  }
}", file="HW1.1.12.txt")

HW1.1.12<-jags(data=list(Y=weightVals,stdev=sd(weightVals)),
  parameters.to.save = "m",
  model.file = "HW1.1.12.txt",
  n.iter=00000,
  n.burnin=10000
)
round(HW1.1.12$BUGSoutput$summary,3)
```

With the vague prior, the results are more similar to the frequentist sampling (mean=`r round(HW1.1.12$BUGSoutput$summary["m","mean"],2)`, sd=`r round(HW1.1.12$BUGSoutput$summary["m","sd"],2)`) distribution and further from the prior. 

## g. Diagnostics

```{r}
gg.1<-ggs(as.mcmc(HW1.1.1))
gg.1<-filter(gg.1,Parameter=="m")
ggs_traceplot(gg.1) +ggtitle("With informative prior")
ggs_density(gg.1)+ggtitle("With informative prior")
```


```{r}
gg.2<-ggs(as.mcmc(HW1.1.12))
gg.2<-filter(gg.2,Parameter=="m")
ggs_traceplot(gg.2)+ggtitle("With vague prior")
ggs_density(gg.2)+ggtitle("With vague prior")
```

Both look converged, based on the low values of Rhat (near 1), high values of n.eff (>3000), smooth density plot, and no pattern in the traceplot.  

# 2. Binomial and negative binomial

## a. Binomial model 

- Data: x= 5 mortalities out of 17

- Parameter: Mortality rate r

- Likelihood:  x~binomial(r,17)

- Prior:  r~uniform(0,1)

## b. Jags MCMC fit

```{r}
write("model{
  x~dbin(r,17)
  r~dunif(0,1)
}",
  file="binom.txt")
bin1<-jags(data=list(x=5),
  parameters.to.save = c("r"),
  model.file = "binom.txt",
  n.iter=100000,
  n.burnin=10000)
print(bin1)
ggb<-ggs(as.mcmc(bin1))
ggb<-filter(ggb,Parameter=="r")
ggs_traceplot(ggb)+ggtitle("Binomial")
ggs_density(ggb)+ggtitle("Binoimal")
```

The mortality rate is `r round(bin1$BUGSoutput$summary["r","mean"],2)` with a credible interval of `r round(bin1$BUGSoutput$summary["r","2.5%"],2)` to `r round(bin1$BUGSoutput$summary["r","97.5%"],2)`. 


## c. Negative binomial model 

- Data: x= 12 survivors before 5 mortalities

- Parameter: Mortality rate r

- Likelihood:  x~NegativeBinomial(r,5)

- Prior:  r~uniform(0,1)

## b. Jags MCMC fit

```{r}
write("model{
  x~dnegbin(r,5)
  r~dunif(0,1)
}",
  file="negbinom.txt")
negbin1<-jags(data=list(x=12),
  parameters.to.save = c("r"),
  model.file = "negbinom.txt",
  n.iter=200000,
  n.burnin=10000)
print(negbin1)
ggnb<-ggs(as.mcmc(negbin1))
ggnb<-filter(ggnb,Parameter=="r")
ggs_traceplot(ggnb)+ggtitle("Negative binomial")
ggs_density(ggnb)+ggtitle("Negative binomial")
```

The mortality rate is `r round(negbin1$BUGSoutput$summary["r","mean"],2)` with a credible interval of `r round(negbin1$BUGSoutput$summary["r","2.5%"],2)` to `r round(negbin1$BUGSoutput$summary["r","97.5%"],2)`. 

## e. Comparison

The negative binomial and binomial estimates are the same except for MCMC sampling error, as expected under the likelihood principal. 
