---
title: "HW 9"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

```{r, message=FALSE}
library(R2jags)
library(tidyverse)
library(ggmcmc)
library(gridExtra)
library(loo)
library(kableExtra)
theme_set(theme_bw())
```

## 1. Croaker growth

## a Plot

```{r}
croaker<-read.csv("croakerF.csv")

ggplot(croaker,aes(x=Age,y=ObsL))+geom_point() 
```

It is hard to say whether normal or lorngormal is best. The residual variance seems to be different in different years with no obvious pattern. 

## b Lognormal

```{r}
write("model 
  {
  for (i in 1:N){ 
    PredL[i] <- L1 +(L2-L1)* (1 - exp(- K* (Age[i] -Age1)))/(1 - exp(- K* (Age2 -Age1)))  # model prediction  
    logPredL[i] <- log(PredL[i])   # log-transformation of pred.value
    ObsL[i] ~ dlnorm(logPredL[i], tau)   # lognormal likelihood
    logObsL[i] <-log(ObsL[i])       # log transfomration of observed value
    resid[i] <- logObsL[i]-logPredL[i]  # residuals
    LL[i]<- -log(1/sqrt(tau))-0.5*log(2*3.14159)-0.5*(logObsL[i]-logPredL[i])^2*tau-logObsL[i]  }
  #priors specification
  K ~ dunif(0,4)
  L1 ~ dunif(10,2000)   
  L2 ~ dunif(10,2000) 
  tau~dgamma(0.001,0.001)
  Age1<-1
  Age2<-10
  #Derived parameters
    Linf<- (L2-L1*exp(-K*(Age2-Age1)))/(1-exp(-K*(Age2-Age1)))
    dev<- -2*sum(LL[])
}
",file="VonBertLnormSchnute.txt")

dat1<-list(Age=croaker$Age,ObsL=croaker$ObsL,N=nrow(croaker))
init1<-list(tau=1,L1=100,L2=1000,K=0.5)
init2<-list(tau=2,L1=200,L2=500,K=0.7)

res1bjags<-jags(dat1,list(init1,init2),model.file="VonBertLnormSchnute.txt",
              parameters.to.save=c("K","L1","L2","Linf","tau","logPredL","PredL",
                                   "resid","LL","dev"),
              n.chains=2,n.iter=11000,n.burnin=1000,n.thin=1)

resb<-res1bjags$BUGSoutput
range(resb$summary[,"n.eff"])
range(resb$summary[,"Rhat"])
resb$summary[c("L1","L2","K","tau","Linf","dev","deviance"),] %>%
  kable(format="html",digits=2)

#PLot residuals
n<-nrow(croaker)
residrows<-paste0("resid[",1:n,"]")
meanrows<-paste0("logPredL[",1:n,"]")
dfcheck<-data.frame(Predicted=resb$summary[meanrows,"mean"],
                    Residual=resb$summar[residrows,"mean"])
g1<-ggplot(dfcheck)+geom_point(aes(x=Predicted,y=Residual))+
  geom_abline(intercept=0,slope=0)+ggtitle("Residuals")
g2<-ggplot(dfcheck,aes(sample=Residual))+geom_qq()+geom_qq_line()+
  ggtitle("QQNormal of Residuals")
grid.arrange(g1,g2,ncol=2)

```

The residuals are plotted on the log scale for the lognormal. There may be a slight deviation from normality in the qqnormal plot but the residual variance looks constant, which is good. 

## c. Normal

```{r}
write("model {
  for (i in 1:N){ 
   PredL[i] <- L1 +(L2-L1)* (1 - exp(- K* (Age[i] -Age1)))/(1 - exp(- K*(Age2 -Age1))) 	#  model prediction  
   ObsL[i] ~ dnorm(PredL[i], tau)  	#Normal likelihood
   resid[i] <- ObsL[i] - PredL[i]  # residuals
   LL[i]<- -log(1/sqrt(tau))-0.5*log(2*3.14159)-0.5*(ObsL[i]-PredL[i])^2*tau
}
 #priors specification
 K ~ dunif(0,4)
 Age1<-1
 Age2<-10
 L1~dunif(10,2000) 	
 L2~dunif(10,2000) 	
 tau~dgamma(0.001,0.001)
 #Derived parameters
 Linf<- (L2-L1*exp(-K*(Age2-Age1)))/(1-exp(-K*(Age2-Age1)))
 dev<- -2*sum(LL[])
}",file="VonBertNormalSchnute.txt")

res1cjags<-jags(dat1,list(init1,init2),model.file="VonBertNormalSchnute.txt",
              parameters.to.save=c("K","L1","L2","Linf","tau","LL","PredL","resid","dev"),
              n.chains=2,n.iter=11000,n.burnin=1000,n.thin=1)
resc<-res1cjags$BUGSoutput
range(resc$summary[,"n.eff"])
range(resc$summary[,"Rhat"])
resc$summary[c("L1","L2","K","tau","Linf","dev","deviance"),]%>%
  kable(format="html",digits=2)

#Plot residuals
n<-nrow(croaker)
residrows<-paste0("resid[",1:n,"]")
meanrows<-paste0("PredL[",1:n,"]")
dfcheck<-data.frame(Predicted=resc$summary[meanrows,"mean"],
                    Residual=resc$summary[residrows,"mean"])
g1<-ggplot(dfcheck)+geom_point(aes(x=Predicted,y=Residual))+
  geom_abline(intercept=0,slope=0)+ggtitle("Residuals")
g2<-ggplot(dfcheck,aes(sample=Residual))+
  geom_qq()+geom_qq_line()+ggtitle("QQNormal of Residuals")
grid.arrange(g1,g2,ncol=2)
```

The normal residuals show an increasing trend in variance, which implies that the lognormal might be better. The qqnormal plot also shows some deviation from normality. Note that the parameters (L1,L2,K) are similar between the two models, but the precision is not comparable because it is on the log scale for lognormal but not normal. 

## d

```{r}
LLrow<-paste0("LL[",1:n,"]")
LLb<-res1bjags$BUGSoutput$sims.matrix[,LLrow]
LLc<-res1cjags$BUGSoutput$sims.matrix[,LLrow]
waic(LLb)
waic(LLc)
         
```

The WAIC values are nearly identical with a deltaWAIC of `r waic(LLc)$estimate[3,1] - waic(LLb)$estimate[3,1]`

## e. Ageing error

```{r}
write("model   {
  for (i in 1:N){ 
   AgeX[i]~dunif(1,10)
   Age[i]~dnorm(AgeX[i],4)
   PredL[i] <- L1 +(L2-L1)* (1 - exp(- K* (AgeX[i] -Age1)))/(1 - exp(- K* (Age2 -Age1)))  # model prediction  
    logPredL[i] <- log(PredL[i])   # log-transformation of pred.value
    ObsL[i] ~ dlnorm(logPredL[i], tau)   # lognormal likelihood
    logObsL[i] <-log(ObsL[i])       # log transfomration of observed value
    resid[i] <- logObsL[i]-logPredL[i]  # residuals
  }  
   #priors specification
  K ~ dunif(0,4)
  L1 ~ dunif(10,2000)   
  L2 ~ dunif(10,2000) 
  tau~dgamma(0.001,0.001)
  Age1<-1
  Age2<-10
  #Derived parameters
    Linf<- (L2-L1*exp(-K*(Age2-Age1)))/(1-exp(-K*(Age2-Age1)))
}
",file="VonBertLnormSchnuteAgeing.txt")

init1<-list(tau=1,L1=c(100),L2=c(2000),K=c(0.5),AgeX=dat1$Age)
init2<-list(tau=2,L1=c(200),L2=c(1000),K=c(0.3),AgeX=dat1$Age)

res1ejags<-jags(dat1,list(init1,init2),model.file="VonBertLnormSchnuteAgeing.txt",
              parameters.to.save=c("K","L1","L2","Linf","tau","AgeX","PredL"),
              n.chains=2,n.iter=11000,n.burnin=1000,n.thin=1)
rese<-res1ejags$BUGSoutput
range(rese$summary[,"n.eff"])
range(rese$summary[,"Rhat"])
rese$summary[c("L1","L2","K","Linf","tau"),]%>%
  kable(format="html",digits=2)
```
The ageing error doesn't change the parameter values very much, so it probably isn't necessary. 

## Marlin ZIP and delta-lognormal

```{r}
marlin<-read.csv("marlin.csv")
summary(marlin)
marlin$YearNum<-marlin$Year-1994

write("model  {
  for(i in 1:5) {
   a[i]~dnorm(0,1.0E-6)
   alpha[i]~ dnorm(0, 1.0E-6)
  }
  for(i in 1:426)  {
   logit(phi[i])<-a[YearNum[i]]
   z[i]~dbern(phi[i])
   log.mean[i]<-alpha[YearNum[i]] 
   mean.pos[i]<-exp(log.mean[i])
   mean.tot[i]<-mean.pos[i] *  z[i]
   Count[i]~dpois(mean.tot[i])
   resid[i]<-(Count[i]-mean.tot[i])
  }
  for(i in 1:5) {
   predmean[i] <- exp(alpha[i])
   predpos[i]<- ilogit(a[i])
   totalpred[i]<-predmean[i] * predpos[i]
}}
",file="ZIP.txt")

init1<-list(list(alpha=rep(1,5),z=rep(1,nrow(marlin))),
  list(alpha=rep(2,5),z=rep(1,dim(marlin)[1])))

res2a<-jags(marlin,init1,parameters.to.save = c("alpha","a","totalpred"),
             model.file = "ZIP.txt",n.thin=4,n.iter = 110000,n.burnin = 10000,
             n.chains = 2)
#Show convergence
range(res2a$BUGSoutput$summary[,"n.eff"][res2a$BUGSoutput$summary[,"n.eff"]>1])
range(res2a$BUGSoutput$summary[,"Rhat"])
res2a$BUGSoutput$summary[c(paste0("alpha[",1:5,"]"),paste0("a[",1:5,"]")),]%>%
  kable(format="html",digits=2)
ggs7.1b<-ggs(as.mcmc(res2a))
ggs_caterpillar(ggs7.1b,family="totalpred",sort=FALSE)+coord_flip()
```

# 2b Delta lognormal

```{r}
write("model  {
  for(i in 1:5) {
   a[i]~ dnorm(0, 1.0E-6) 
   alpha[i]~ dnorm(0, 1.0E-6)
 }
  prec~dgamma(0.01,0.001)
  for(i in 1:N1)  {
   logit(p[i])<-a[YearNum1[i]] 
   present[i]~dbern(p[i])
  }
  for(i in 1:N2)  {
   mean[i]<-alpha[YearNum2[i]] 
   count[i]~dlnorm(mean[i],prec)
  }
   for(i in 1:5) {
    mean.lnorm[i] <- exp(alpha[i]+1/(2*prec))
    logit(predp[i])<-a[i]
    total.mean[i] <-predp[i]* mean.lnorm[i]
   }
}",file="DeltaLognormal.txt")

marlin$present<-ifelse(marlin$Count>0,1,0)
deltadat1<-list(N1=nrow(marlin),N2=sum(marlin$present),
             present=marlin$present,
             count=marlin$Count[marlin$present==1],
             YearNum1=marlin$YearNum,
             YearNum2=marlin$YearNum[marlin$present==1])
summary(deltadat1)

init1<-list(list(prec=10,alpha=rep(0,5),a=rep(2,5)),
  list(prec=1,a=rep(1,5),alpha=rep(1,5)))

resres2b<-jags(deltadat1,inits=init1,
  parameters.to.save = c("alpha","a","total.mean","prec"),
             model.file = "DeltaLognormal.txt",
              n.thin=1,n.iter = 11000,
              n.burnin = 1000,
             n.chains = 2)
#Show convergence
range(resres2b$BUGSoutput$summary[,"n.eff"][resres2b$BUGSoutput$summary[,"n.eff"]>1])
range(resres2b$BUGSoutput$summary[,"Rhat"])

# Parameter summaries
resres2b$BUGSoutput$summary[c("prec",paste0("alpha[",1:5,"]"),paste0("a[",1:5,"]")),]%>%
  kable(format="html",digits=2)

# Estimated mean
ggsres2b<-ggs(as.mcmc(resres2b))
ggs_caterpillar(ggsres2b,family="total.mean",sort=FALSE)+coord_flip()
```
The trend looks similar to the others. 

# d. Comparison
The delta lognormal gave similar results and was much faster, but it can't be compared to the other models using information critera because of the hybrid model structure.    
