---
title: "lecture 16_other R libraries"
author: "Beth Babcock"
date: "2024-03-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r,message=FALSE}
library(tidyverse)
library(ggmcmc)
library(BayesFactor)
library(rstan)
library(rstanarm)
library(shinystan)
library(lme4)
library(loo)
library(brms)
library(here)
theme_set(theme_bw())
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```


## rstanarm

```{r}
sleepstudy$Reaction<-sleepstudy$Reaction/100
rstanSleep<-stan_glm(Reaction~Days,data=sleepstudy)
summary(rstanSleep)

#Getting the generated rstan code is less useful than it sounds.
#get_stanmodel(rstanSleep$stanfit)
#To see priors
prior_summary(rstanSleep)
#?prior_summary
#?stan_glm

#to specify priors, both of these are the same as the default
rstanSleep2<-stan_glm(Reaction~Days,data=sleepstudy,
  prior_aux=rstanarm::exponential(1,autoscale=TRUE))
rstanSleep3<-stan_glm(Reaction~Days,data=sleepstudy,
  prior_aux=rstanarm::exponential(1/sd(sleepstudy$Reaction), autoscale=FALSE))
prior_summary(rstanSleep2)
prior_summary(rstanSleep3)

```
You can see that the scaled exponential prior on the standard deviation has a mean equal to the sd of the Y data.  

```{r}
stan_trace(rstanSleep)
waic(rstanSleep)
```

To run shinystan and look at all the diagnostics

```{r}
#launch_shinystan(rstanSleep)

```

To get the replicated Y data without having to add the calculation in the code. 

```{r}
y_rep <- posterior_predict(rstanSleep)
dim(y_rep)

```

Can also make the same posterior predictive density plots with pp_check

```{r}
pp_check(rstanSleep)
```


## Mixed models 

```{r}
#?stan_glmer

rstanSleep2<-stan_glmer(Reaction~Days+(Days|Subject),data=sleepstudy)
#launch_shinystan(rstanSleep2)
```

Poisson regression example

```{r}
Epil<-INLA::Epil %>% mutate(
  CTrt = Trt-mean(Trt),
  ClBase4 = log(Base/4)-mean(log(Base/4)),
  CV4 = V4-mean(V4),
  ClAge = log(Age)-mean(log(Age)),
  CBT = CTrt*ClBase4)
summary(Epil)

rsEpil<-stan_glmer(y ~ ClBase4 + CTrt + CBT+ ClAge + CV4 +(1|Ind) + (1|rand), 
                   data=Epil,
                   family="poisson")
#launch_shinystan(rsEpil)
prior_summary(rsEpil)

```

Logistic regression example

```{r}
lr<-read.csv(here("data", "lr.csv"))

rslr<-stan_glm(lr~numrock+volcanic+rockcell+nearslope+nearslope^2,
               data=lr,
               family="binomial")
#launch_shinystan(rslr)
```

Anova example

```{r}
starling<-read.csv(here("data", "starling.csv")) %>%
  mutate(location=factor(location),
         period=factor(period))
rsStarling<-stan_glm(mass~period*location,
                     data=starling)
rsStarling2<-stan_glm(mass~period+location,
                     data=starling)
#launch_shinystan(rsStarling)
```

## brms

https://paul-buerkner.github.io/brms/

```{r}

brmsEpil<-brm(y ~ ClBase4 + CTrt + CBT+ ClAge + CV4 +(1|Ind) + (1|rand), 
                   data=Epil,
                   family="poisson")

brmsEpil1<-brm(y ~ ClBase4 + CTrt + CBT+ ClAge + CV4 +(1|Ind) , 
                   data=Epil,
                   family="poisson")
                   
#launch_shinystan(brmsEpil)

waic(brmsEpil,brmsEpil1)
loo(brmsEpil,brmsEpil1)
```

Useful brms functions

```{r}
make_stancode(y ~ ClBase4 + CTrt + CBT+ ClAge + CV4 +(1|Ind) , 
                   data=Epil,
                   family="poisson")
get_prior(y ~ ClBase4 + CTrt + CBT+ ClAge + CV4 +(1|Ind) , 
                   data=Epil,
                   family="poisson")
prior_summary(brmsEpil)

brmsEpil2<-brm(y ~ ClBase4 + CTrt + CBT+ ClAge + CV4 +(1|Ind) , 
                   data=Epil,
                   family="poisson",
                   prior = prior(exponential(1),class="sd"))
prior_summary(brmsEpil2)
```

## Using BayesFactor for anova

```{r}
ggplot(starling,aes(x=location,y=mass,color=period))+stat_summary(size=1)

#Compare Bayesian models with WAIC, 
sumtab<-data.frame(Model=c("Full","No int"),
  WAIC=c(waic(rsStarling)$waic,waic(rsStarling2)$waic) )  %>% 
  mutate(DeltaWAIC=WAIC-min(WAIC),
  WAIC.wt=exp(-DeltaWAIC/2)/sum(exp(-DeltaWAIC/2)))
sumtab
```

## Compute BayesFactor

```{r}
?anovaBF
bf1 <- anovaBF(mass~location*period,data=starling)
bf1
bf1[3]
bf1[4]/bf1[3]  #compare full to main effect only
bf1[3]/bf1[4]  #without interaction is 4 times better
extractBF(bf1)$bf  #get the number
plot(bf1)
plot(bf1[2:4]/bf1[1])
```

Ratio is 4.4, so model odds are 3.8 times higher for the model without interactions
BF >1 is better.

Model weights/probabilities can be calculated, assuming equal prior probabilities,

```{r}
BFratio<-extractBF(bf1)$bf[3]/extractBF(bf1)$bf[4]
BFratio
sumtab$BFprob<-c(1/(BFratio+1),1/(1/BFratio+1))
sumtab
```
Diagnostics

```{r}
bf1NoInt<-ggs(as.mcmc(posterior(bf1,iterations=5000, index=3)))
bf1Full<-ggs(as.mcmc(posterior(bf1,iterations=5000, index=4)))
ggs_caterpillar(bf1NoInt)
ggs_caterpillar(bf1Full)
```

## Regression example with BayesFactor

```{r}

#?lmBF
BFReaction<-lmBF(Reaction~Days,data=sleepstudy)
BFReaction
```
