---
title: "Lecture 5"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

# R code for JAGS examples and plots from lecture 5 

```{r,message=FALSE,error=FALSE}
library(tidyverse)
library(R2jags)
library(ggmcmc)
library(here)
theme_set(theme_bw())
```

## Box 3.2 Estimating mean and variance from normal data

```{r}
write("
model
{
  mean ~ dnorm(0, 1.0E-6)  	# mean diameter of trees in the remnant (uninformative prior)
  Var ~ dlnorm(0.0, 1.0E-6)  # variance of tree diameters (uninformative prior)
 # the following replace the above when using informative priors
 # mean ~ dnorm(53, 0.04)  # prior with mean 53 and sd 5 (prec=1/(5*5))
 # var ~ dlnorm(4.75, 3.05)  # informative prior for variance
  prec <- 1/Var							# converts variance to precision
  for (i in 1:10)  					# for each of the ten trees
  {
    Y[i] ~ dnorm(mean, prec)  # tree diameter drawn from normal distribution
  }
}
",here("JAGS_mods","model3.2var.txt"))

Data<- list(Y=c(42, 43, 58, 70, 47, 51, 85, 63, 58, 46)) # site 2 sample

InitialValues<- list(list(Var=100, mean=100),
  list(Var=200,mean=50))

model3.2var.res<-jags(data=Data,
  inits=InitialValues,
  parameters.to.save = c("mean","Var","prec"),
  n.chains = 2,
  n.burnin = 1000,
  n.iter = 20000,
  n.thin = 1,
  model.file = here("JAGS_mods","model3.2var.txt"))

print(model3.2var.res)
gg3.2var<-ggs(as.mcmc(model3.2var.res))  #Make the JAGS output an MCMC object and make the MCMC object a ggs object for plotting

#note that in the ggs_pairs the diagonal plots are marginal distributions,
#the lower subdiagonal is the joint distributions, and the upper subdiagonal
#is the correlation between the 2 parameters.

ggs_pairs(filter(gg3.2var,Parameter %in% c("mean","Var")))  
ggs_pairs(filter(gg3.2var,Parameter %in% c("mean","Var")),lower=list(continuous="density"))
ggs_density(gg3.2var)  #Density plot
ggs_traceplot(gg3.2var)  # Trace plot
```

## Alternative mean priors

```{r}
write("
model
{
  mean ~ dnorm(50, 0.0001)  	# mean with vague prior
  Var ~ dlnorm(0.0, 1.0E-6)  # variance of tree diameters (uninformative prior)
  prec <- 1/Var							# converts variance to precision
  for (i in 1:10)  					# for each of the ten trees
  {
    Y[i] ~ dnorm(mean, prec)  # tree diameter drawn from normal distribution
  }
}
",here("JAGS_mods","model3.2var2.txt"))

model3.2var2.res<-jags(data=Data,
  inits=InitialValues,
  parameters.to.save = c("mean","Var","prec"),
  n.chains = 2,
  n.burnin = 1000,
  n.iter = 20000,
  n.thin = 1,
  model.file = here("JAGS_mods","model3.2var2.txt"))

#compare results
gg3.2var2<-ggs(as.mcmc(model3.2var2.res)) 
gg3.2<-bind_rows(list(`1000`=filter(gg3.2var,Parameter=="mean"),
  `100`=filter(gg3.2var2,Parameter=="mean")),.id="Prior mean sd") %>%
  mutate(Chain=factor(Chain))
ggplot(gg3.2,aes(x=value,fill=Chain,color=Chain))+
  geom_density(alpha=0.5)+
  facet_wrap(`Prior mean sd`~.,ncol=1)
bind_rows(list(`1000`=round(model3.2var.res$BUGSoutput$summary["mean",],2),
  `100`=round(model3.2var2.res$BUGSoutput$summary["mean",],2)),.id="Prior mean sd")

```

## Alternative variance priors.

```{r}
n=100
priors<-data.frame(prior=rep(c("lognormal on variance",
  "uniform  on standard deviation",
  "uniform on log(sd)", 
  "gamma on precision",
  "exponential on sd"),each=n)) %>% 
mutate(sd=rep(seq(0.01,10,length=n),5),
  var=sd^2,
  prec=1/var,
  logsd=log(sd)) %>%
mutate(Probability=case_when(prior=="lognormal on variance"~dlnorm(var,0,1),
  prior=="uniform  on standard deviation"~dunif(sd,0,10),
  prior=="uniform on log(sd)"~dexp(sd,1/log(5)), 
  prior=="gamma on precision"~dgamma(prec,2,1),
  prior=="exponential on sd"~dexp(sd,1),
  TRUE~NA))

ggplot(priors,aes(x=sd,y=Probability,color=prior))+
  geom_line(linewidth=1)+
  ggtitle("Priors for standard deviation")

```

## Running the model will several alternative variance priors

Change the comments to change the prior set up. 

```{r}
write("model
{
  mean ~ dnorm(100, 1E-6)  #prior for mean	
  #Prior on precision
#  prec ~ dgamma(0.01,0.001)   
#  Var <- 1/prec
#  sd <- sqrt(Var)
  #Prior on sd
  sd~dexp(0.1)
#  sd~dunif(0,100)   
#  sd~dunif(0,10)  #Prior that is too constrained and biases answer
  Var <- sd *sd
  prec <- 1/Var
  #likelihood
  for (i in 1:10)  		 	
  {
    Y[i] ~ dnorm(mean, prec)
  }
}
", file=here("JAGS_mods","model3.2alt.txt"))

#If you initialize, must be on estimated parameter
model3.2alt.res<-jags(data=Data,
#   inits=list(list(prec=1),list(prec=2)),
  inits=list(list(sd=2),list(sd=1)),  #for sd as the estimated parameter
#  inits=list(list(Var=2),list(Var=1)),  #for Var as the estimated parameter
  parameters.to.save = c("mean","Var","prec","sd"),
  n.chains = 2,
  n.burnin = 1000,
  n.iter = 20000,
  n.thin = 1,
  model.file = here("JAGS_mods","model3.2alt.txt"))

print(model3.2alt.res)
gg3.2alt<-ggs(as.mcmc(model3.2alt.res))  #Make the JAGS output an MCMC object and make the MCMC object a ggs object for plotting
gg3.2alt<-filter(gg3.2alt,Parameter %in% c("sd"))
ggs_density(gg3.2alt)  #Density plot
```

# Box 3.7 Mean difference and using step for probabilities

##(KIND OF LIKE A PAIRED-T.TEST)

```{r}
write("model{
  vmeandiff ~ dnorm(0, 1.0E-6)					# uninformative prior for mean vert. diff.
  prec ~ dgamma(0.001, 0.001) 					# uninf. prior for precision of vert. diff.
  probpos<-step(vmeandiff)
  for (i in 1:17)  											# for each of the 17 spiders
  {
    VertDiff[i] ~ dnorm(vmeandiff, prec)  # observed diff. drawn from a normal dist'n
  }
}",file=here("JAGS_mods","model3.7.txt"))

InitialValues<-list(list(vmeandiff=0, prec=100),list(vmeandiff=10, prec=200))
Data<-list(VertDiff=c(-220,-120,-80	,-130,	-10	,-60	,-40	,60	,-10	,40	,-30	,30	,30	,110	,
90	,70	,-80	))

model3.7.res<-jags(data=Data,
  inits = InitialValues,
  parameters.to.save = c("vmeandiff","prec","probpos"),
  n.chains = 2,
  n.burnin = 1000,
  n.iter = 20000,
  n.thin = 1,
  model.file=here("JAGS_mods","model3.7.txt"))

round(model3.7.res$BUGSoutput$summary,4)
```

## Looking at individual mcmc draws.

```{r}
dim(model3.7.res$BUGSoutput$sims.array)
dim(model3.7.res$BUGSoutput$sims.matrix)
ggs3.7<-ggs(as.mcmc(model3.7.res))
head(ggs3.7)
ggs_pairs(filter(ggs3.7,Parameter %in% c("vmeandiff","probpos")))
ggs_density(filter(ggs3.7,Parameter %in% c("vmeandiff")))
ggs3.7 %>% filter(Parameter == "vmeandiff") %>%
  summarize(ProbPos=mean(value>0))
#or
vmeandiff<-model3.7.res$BUGSoutput$sims.matrix[,"vmeandiff"]
length(vmeandiff[vmeandiff>0])/length(vmeandiff)

```
# Sample size for given posterior sd

```{r}
sampleSize<-data.frame(priorsd=c(1000,22.8)) %>%
  mutate(required.n=94.8^2*(1/100-1/priorsd^2))
sampleSize
```

# Box 3.4 Poisson mean

```{r}
write("model{
  for (i in 1:10)					# for each of the ten quadrats
  {
     y[i] ~ dpois(m)			# number of trees drawn from Poisson with mean m
  }

  m ~ dlnorm(0.0, 1.0E-6)  # uninformative prior for mean trees per quadrat
}",here("JAGS_mods","model3.4.txt"))

Data<-list(y=c(6,0,1,2,1,7,1,5,2,0))

InitialValues<-list(list(m=5),list(m=4))

model3.4.res<-jags(data=Data,
  inits = InitialValues,
  parameters.to.save = c("m"),
  n.chains = 2,
  n.burnin = 1000,
  n.iter = 20000,
  n.thin = 1,
  model.file=here("JAGS_mods","model3.4.txt"))

model3.4.res$BUGSoutput
ggs3.4<-filter(ggs(as.mcmc(model3.4.res)),Parameter=="m")
ggs_density(ggs3.4)

#Fit compared to data
ggplot(data.frame(Trees=Data$y),aes(x=Trees))+geom_histogram(bins=8,fill="lightblue",color="darkblue")+
  geom_line(data=data.frame(Trees=0:7,Count=length(Data$y)*dpois(0:7,2.5)),aes(x=Trees,y=Count),lwd=2)
```
  
